
\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{natbib}
\usepackage{url}

\newcommand{\inner}[1]{\langle #1 \rangle}
\newcommand{\set}[1]{\{\, #1 \,\}}
\newcommand{\abs}[1]{\lvert #1 \rvert}
\newcommand{\fatdot}{\,\cdot\,}

\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\logit}{logit}

\newcommand{\real}{\mathbb{R}}

\begin{document}

  \vspace*{0.9375in}
  \begin{center}
    {\bfseries A Philosophical Look at Aster Models} \\
    By \\
    Charles J. Geyer \\
    Technical Report No.~676 \\
    School of Statistics \\
    \today
  \end{center}
  \thispagestyle{empty}
  \cleardoublepage
  \setcounter{page}{1}
  \thispagestyle{empty}

\section{Introduction} \label{sec:intro}

Aster models \citep*{gws,aster2} are statistical models designed especially
for life history analysis of plants and animals.

Aster models are parametric statistical models.  In some senses they are
quite ordinary statistical models.  The theory of aster models is just
the theory of likelihood inference \citep{fisher,severini} and
the theory of exponential families of distributions
\citep{barndorff,brown,thesis}.
Given an aster model and data, we use the method
of maximum likelihood for parameter estimation; we use likelihood ratio
tests for model comparison; and we use confidence intervals based on
the asymptotic distribution of the maximum likelihood estimator (MLE).
Details are given in \citet {gws};
see also Section~\ref{sec:efficiency} below.

The theory mentioned in the preceding paragraph is known to all statisticians,
covered, albeit not completely rigorously, in first-year graduate courses in
theoretical statistics and covered again, this time completely rigorously,
in second-year graduate courses in theoretical statistics.  Many scientists
in a wide variety of disciplines take a first-year graduate course in
theoretical statistics, so this theory is very widely known.

Aster models are regression models.  In some senses they are
quite like linear models (LM), which include multiple linear regression
and analysis of variance (explained in many textbooks, for example,
\citealp{weisberg} and \citealp{oehlert}), and generalized
linear models (GLM) \citep{mccullagh}.  In regression models the objective
is to model the conditional distribution of the response vector $y$ given
predictor data $x$ (the structure of which does not matter).
Aster models have the same objective.  In LM and GLM, conditional on $x$
the components of $y$ are assumed to be stochastically
independent and to have marginal distributions from the same family
e.~g., all normal, all Bernoulli, or all Poisson.
In aster models, in contrast, conditional on $x$
the components of $y$ are allowed to be stochastically dependent and are
allowed to have conditional distributions from different families
e.~g., some normal, some Bernoulli, and some Poisson.

LM and GLM are known to all statisticians,
covered, albeit not at all rigorously, in undergraduate statistics courses
and covered again, this time much more rigorously,
in first-year graduate courses in applied statistics (proofs
are left for theoretical statistics courses).  Many scientists
in a wide variety of disciplines take these courses, so these models
are very widely known.  Some courses teach ways of thinking about LM
that do not generalize to GLM and ways of thinking about GLM
that do not generalize to aster models.\footnote{They may imply that
the word ``regression'' only applies to a subset of LM (multiple regression but
not analysis of variance) not to GLM or aster models.  This, of course, cannot
be because the terms ``logistic regression'' and ``Poisson regression'' for
certain GLM are well established, as is the term ``nonparametric regression''
for models that are neither LM nor GLM.
Most courses about LM teach ``response is mean + error''
$$
   y = \mu + e
$$
but this does not generalize even to GLM.  See also Sections~\ref{sec:submodel}
and~\ref{sec:sufficient}.}  Thus people who have had courses covering
LM and GLM may have to unlearn certain things they learned about LM and GLM
and learn more general concepts to understand aster models.

Aster models are graphical models \citep{lauritzen}, but
a very special case thereof.  The only idea aster models take from graphical
model theory is the fundamental idea of representing conditional distribution
relationships between random variables by graphs.  None of the mathematical
theory of graphical models is used in aster model theory.  Although there
are a number of recent textbooks covering graphical models, many statistics
departments do not teach courses using them.  Thus the theory of graphical
models is not widely known (few statisticians are expert in it),
so it is fortunate aster models need so little of the theory
of graphical models.

The ``linear'' in LM comes from the fact that means are modeled
linearly\footnote{The term ``linear'' is used here in the technical
mathematical sense, as in ``linear equations'' and ``linear algebra''
and has no connection with the vague metaphorical sense popular
in postmodernism, as in ``linear thinking.''}
\begin{equation} \label{eq:lm}
   \mu = M \beta,
\end{equation}
where $\mu = E(y)$ is the mean of the response vector,\footnote{Strictly
speaking, we should write $\mu = E(y \mid x)$, a conditional
expectation, instead of $\mu = E(y)$, an unconditional expectation,
because we are conditioning on the predictor data $x$, but we follow
common practice of not explicitly indicating this in the notation.  The fact
that $\mu$ depends on $x$ is implicit in \eqref{eq:lm} because $M$ is allowed
to be an arbitrary function of $x$.}
where $M$ is a known matrix,
called the \emph{model matrix}, which may be an arbitrary function of
predictor data $x$, where $\beta$ is an unknown parameter to be estimated,
and where $M \beta$ is multiplication of a matrix $M$ and a vector $\beta$
yielding a vector $\mu$.
Usually, the dimension of $\beta$ is much less than the dimension of $\mu$
(which is the dimension of $y$).  This ``dimension reduction'' in the jargon
allows the explanation of something complicated involving many parameters
(the components of $\mu$) by something simple involving fewer parameters
(the components of $\beta$), a ``parsimonious model'' in the jargon.
It also contributes to efficient estimation (Section~\ref{sec:efficiency}
below); the fewer the parameters, the
better the estimation, in general.

The ``linear'' in GLM comes from the fact that parameters other than means
are modeled linearly
\begin{equation} \label{eq:glm}
   \eta = M \beta,
\end{equation}
where $\eta$ is a new parameter vector, called the \emph{linear predictor},
and $M$ and $\beta$ are as in LM.  The linear predictor is a componentwise
monotone function of means
$$
   \eta_i = g(\mu_i),
$$
where $g$ is a monotone (strictly increasing) function called
the \emph{link function} and where $\eta_i$ and $\mu_i$ denote components
of $\eta$ and $\mu$.  The reason for this monotone change-of-parameter
is that modeling means linearly makes no sense when possible mean values are
constrained.  When $y$ has Bernoulli (zero-or-one-valued)
components, their means satisfy
$$
   0 \le \mu_i \le 1
$$
but linear modeling \eqref{eq:lm} does not respect these constraints.
The link function is often chosen so that the linear predictor is
unconstrained.  For example, in Bernoulli GLM the link function
\begin{equation} \label{eq:logit}
   \eta_i = \logit(\mu_i) = \log \left( \frac{\mu_i}{1 - \mu_i} \right)
\end{equation}
is the default (other link functions are allowed, but this is by far
the most widely used), and one easily verifies that as $\mu_i$ goes from
zero to one, $\eta_i$ goes from minus infinity to plus infinity.
GLM have a weak connection to exponential family theory in that the default
link function is the one that makes the linear predictor the exponential
family canonical parameter, for example, the logit link \eqref{eq:logit}
for Bernoulli GLM and the log link
$$
   \eta_i = \log(\mu_i)
$$
for Poisson GLM (again one easily verifies that as $\mu_i$ goes from zero to
plus infinity, the allowed values for Poisson means, $\eta_i$ goes from
minus infinity to plus infinity).\footnote{Not all exponential family
canonical parameters are unrestricted, see footnote~\ref{foot:restrict}.}

The ``linear'' in aster models comes from the fact that, just as in GLM,
parameters other than means are modeled linearly
\eqref{eq:glm}.\footnote{Actually \citet{gws} describe a more general type
of modeling in which \eqref{eq:glm} is generalized to \label{foot:affine}
\begin{equation} \label{eq:gam}
   \eta = a + M \beta,
\end{equation}
where $a$ is a known vector, called the \emph{offset vector},
and this formula is
also used (rarely) with GLM.  The offset term $a$ is not needed for most
GLM and aster models and we ignore it here.  When \eqref{eq:gam} is used
instead of \eqref{eq:glm} the proper term is ``affine'' rather than ``linear''
for this change-of-parameter.}
In aster
models, unlike GLM, no choice of link function is allowed;
the linear predictor vector $\eta$ is always the exponential
family canonical parameter vector.  Since components of the
response are allowed to be dependent in aster models, there is no simple
relationship between marginal distributions of components of the response
and their joint distribution (unlike LM and GLM where stochastic independence
implies the joint is the product of the marginals).  Thus the componentwise
monotone relationship of linear predictor and means used in GLM makes less sense
for aster models.\footnote{Emily Grosholz (personal communication) points out
that the ``thus'' here is not justified.  There is no logical reason
(apparent to the author) why there
could not, in principle, be an aster model competitor with componentwise
monotone relationship between unconditional mean values and some other
parameters.  Such a competitor could not have all properties of aster models;
see Section~\ref{sec:discuss} below.}
Instead, exponential family theory implies that the mean
vector $\mu$ is a multivariate strictly monotone function of $\eta$
\citep[p.~121]{barndorff}.
The notion of a multivariate monotone function can be characterized in several
different ways \citep[Chapter~12]{raw}; here is one.
Suppose $\eta_1$ and $\eta_2$ are two different allowed values of the linear
predictor parameter vector and $\mu_1$ and $\mu_2$ are the corresponding
values of the mean vector.  Then
$$
   \inner{\mu_1 - \mu_2, \eta_1 - \eta_2} > 0,
$$
where $\inner{\fatdot, \fatdot}$ denotes the bilinear form defined by
\begin{equation} \label{eq:bilinear}
   \inner{\mu, \eta} = \sum_{i = 1}^p \mu_i \eta_i,
\end{equation}
$p$ being the dimension of $\mu$ and $\eta$.
It may be thought that this property is too abstract to be of much use;
it is certainly not taught in any of the theoretical or applied statistics
courses mentioned above.  But it is a very strong property that enables
important arguments about aster modeling
\citep[appendix on monotonicity; see also Section~\ref{sec:moron}
below]{aster3}.

\section{Key Ideas behind Aster Models}

\subsection{Graphical Models}

Prototypical data for an aster model can be pictured as shown
in \eqref{gra:one}, which represents the data for one individual, which
comprises eight measurements (components of the response vector $y$),
denoted $y_1$, $\ldots$, $y_8$.
\begin{equation} \label{gra:one}
\begin{CD}
   1
   @>\text{Ber}>>
   y_1
   @>\text{Ber}>>
   y_2
   @>\text{Ber}>>
   y_3
   @>\text{Ber}>>
   y_4
   \\
   @.
   @VV\text{Poi}V
   @VV\text{Poi}V
   @VV\text{Poi}V
   @VV\text{Poi}V
   \\
   \hphantom{1}
   @.
   y_5
   @.
   y_6
   @.
   y_7
   @.
   y_8
\end{CD}
\end{equation}
These data are measurements relating to four time periods.
The variables $y_1$, $\ldots$, $y_4$ are Bernoulli and indicate
survival; $y_i = 1$ indicates that the individual was alive at the beginning
of the $i$-th time period, $y_i = 0$ indicates that the individual was dead
at the beginning of this time period.
The variables $y_5$, $\ldots$, $y_8$ are nonnegative-integer-valued
and count offspring; $y_i$ is
the number of offspring the individual had in time period $i - 4$.

The arrows represent stochastic dependence (the idea taken from the theory of
graphical models).  The variable at the head of
an arrow depends on the variable or constant at the tail of the arrow.
For example,
$y_1 = 0$ implies $y_2 = 0$ and $y_5 = 0$ (if the individual is dead
at the beginning of the first time period, then it remains dead at the
beginning of the second time period and cannot have offspring).

We say the variable at the head of an arrow is the \emph{successor} of
the variable or constant at the tail of the arrow (the leftmost arrow has
the constant 1 at its tail).
Conversely, we say the variable or constant at the tail of an arrow is
the \emph{predecessor} of the variable at the head of the arrow.
For example, $y_3$ is the successor of $y_2$ and the predecessor of $y_4$
and $y_7$.

The graph represents stochastic dependence much more precisely than we have
explained so far.  It represents a factorization of the joint distribution
of these random variables as a product of conditional distributions
\begin{equation} \label{eq:factorize}
   f(y_1, \ldots, y_8) = \prod_{j = 1}^8 f(y_j \mid y_{p(j)}),
\end{equation}
where $p$ is the predecessor function: $y_{p(j)}$ denotes the variable
or constant that is the predecessor of $y_j$ (when $y_{p(j)}$ is a constant
the conditional distribution of $y_j$ given $y_{p(j)}$ is, in effect,
unconditional).

% In order for predecessors to be given by a function $J \to J \cup F$
% (when they are already defined by a graph) it must be the case that each
% variable has at exactly one predecessor (which may be either a variable
% or a constant) and that no constant has a predecessor (since constants
% are not random variables and do not have distributions).
% In order for \eqref{eq:factorize} to hold the graph must be \emph{acyclic}

Aster models are a very special case
of graphical models in which each node of the graph (representing a random
variable or a constant) has at most one predecessor and the graph is
\emph{acyclic}, meaning there is no path of arrows joined head to tail that
goes around in a loop.  Such a graph is called
a \emph{forest} in graph theory.

\citet{gws} formalize this notion as follows.\footnote{\citet{gws} actually
formalize a more general class of aster models in which groups of variables
are allowed to have joint conditional distributions given their predecessors
(which are still single variables).  This generalization has not yet been
implemented in software.  We ignore it in this
technical report.\label{foot:group}}
Let $J$ and $F$ be disjoint
finite sets and let $p : J \to J \cup F$ be a map.  For each $j \in J \cup F$
there is a random variable or a constant denoted $y_j$.  Furthermore, $y_j$
is a random variable when $j \in J$ and a constant when $j \in F$.  Let $y_J$
denote the random variables thought of as a single object and $y_F$ the
constants thought of as a single object.  Then the joint probability
distribution of the random variables factorizes as
\begin{equation} \label{eq:factorize-too}
   f(y_J \mid y_F) = \prod_{j \in J} f(y_j \mid y_{p(j)}).
\end{equation}
Again, these conditional distributions are, in effect, unconditional when
what is ``behind the bar'' is constant.  Thus on the left-hand side, because
$y_F$ is a constant vector, we have, in effect, an unconditional distribution.
And on the right-hand side we have some truly conditional distributions,
when $p(j) \in J$, and some in-effect unconditional distributions,
when $p(j) \in F$.

The factorization \eqref{eq:factorize-too} goes with a graph that has an
arrow $y_{p(j)} \to y_j$ for each $j \in J$.  It
generalizes \eqref{eq:factorize}, which is specific
to the graph \eqref{gra:one}, not only in allowing a general graph but also
in allowing more constants
\begin{equation} \label{gra:general}
\begin{CD}
   y_1
   @>>>
   y_2
   @>>>
   y_3
   @>>>
   y_4
   @>>>
   y_5
   \\
   @.
   @VVV
   @VVV
   @VVV
   @VVV
   \\
   \hphantom{y_1}
   @.
   y_6
   @.
   y_7
   @.
   y_8
   @.
   y_9
\\
   y_{10}
   @>>>
   y_{11}
   @>>>
   y_{12}
   @>>>
   y_{13}
   \\
   @.
   @VVV
   @VVV
   @VVV
   \\
   \hphantom{y_{10}}
   @.
   y_{14}
   @.
   y_{15}
   @.
   y_{16}
\\
   y_{17}
   @>>>
   y_{18}
   @>>>
   y_{19}
   \\
   @.
   @VVV
   @VVV
   \\
   \hphantom{y_{17}}
   @.
   y_{20}
   @.
   y_{21}
\end{CD}
\end{equation}
In \eqref{gra:general} $y_1$, $y_{10}$, and $y_{17}$ are constants,
because they have no predecessors.
Each part of the graph connected by a chain of arrows to one
of these constants is called a \emph{maximal connected component}
of the graph.\footnote{Each maximal connected component of a forest graph
is called a \emph{tree} in graph theory (hence the name ``forest'' for a
collection of trees), but we avoid this biological
terminology, because aster models have biological applications, which
may involve real biological trees.}
% The random variables in different maximal connected components are
% stochastically independent.

Returning to biology, different maximal connected
components represent data on different individuals.  Variables within one
maximal connected component represent different measurements on a single
individual.  As shown in \eqref{gra:general} it is not necessary that all
individuals have the same number of variables connected in the same graphical
structure: in \eqref{gra:general} there are three individuals (maximal
connected components), and each is different from the others.

However, in all applications of aster models that have been done so far
all individuals have the same graphical structure and the current version
of the aster software \citep{package}
encourages users to think this way.  In all papers
about aster models that have been written so far the graphs shown refer
to one individual, and it is assumed that all individuals have the
same graphical structure (in the full graph, never shown, all maximal
connected components are isomorphic subgraphs, and what is shown is
just one of these subgraphs).

% \citet{gws} actually describe a more complex graphical structure than the
% one described here, but that more complex structure is not implemented in
% the current version of the aster software.  What is described in this section
% is aster models users can currently use.

Aster model graphs can be much more complicated than those shown above.
They can be arbitrarily complicated subject to the conditions described
above that make the graph a forest (acyclic and each node has at most one
predecessor).  Here is an example taken from \citet{aster3}
\begin{equation} \label{gra:complicated}
\begin{CD}
   1
   @>\text{Ber}>>
   y_1
   @>\text{Ber}>>
   y_2
   @>\text{Ber}>>
   y_3
   @>\text{Ber}>>
   y_4
   \\
   @.
   @VV\text{Ber}V
   @VV\text{Ber}V
   @VV\text{Ber}V
   @VV\text{Ber}V
   \\
   \hphantom{1}
   @.
   y_5
   @.
   y_6
   @.
   y_7
   @.
   y_8
   \\
   @.
   @VV\text{0-Poi}V
   @VV\text{0-Poi}V
   @VV\text{0-Poi}V
   @VV\text{0-Poi}V
   \\
   \hphantom{1}
   @.
   y_9
   @.
   y_{10}
   @.
   y_{11}
   @.
   y_{12}
   \\
   @.
   @VV\text{Poi}V
   @VV\text{Poi}V
   @VV\text{Poi}V
   @VV\text{Poi}V
   \\
   \hphantom{1}
   @.
   y_{13}
   @.
   y_{14}
   @.
   y_{15}
   @.
   y_{16}
   \\
   @.
   @VV\text{Ber}V
   @VV\text{Ber}V
   @VV\text{Ber}V
   @VV\text{Ber}V
   \\
   \hphantom{1}
   @.
   y_{17}
   @.
   y_{18}
   @.
   y_{19}
   @.
   y_{20}
\end{CD}
\end{equation}

We still have not explained the labels on the arrows in \eqref{gra:one}
and \eqref{gra:complicated}.
That will have to wait until Section~\ref{sec:expo}.

\subsection{Predecessor as Sample Size} \label{sec:sample-size}

The dependence in each conditional distribution of $y_j$ given
$y_{p(j)}$ in \eqref{eq:factorize-too} is of a very specific form:
the predecessor plays the role of
sample size for the successor.  This means that $y_j$ is the sum of $y_{p(j)}$
independent and identically distributed (IID) random variables.
By convention, a sum having zero terms is defined to be zero.
Hence $y_{p(j)} = 0$ implies $y_j = 0$ with probability one.

The ``predecessor is sample size'' property has the following consequence.
Define\footnote{These definitions differ from those in \citet{gws} and the
documentation for the R package \texttt{aster} \citep{package}.  They used
$\tau_j$ instead of $\mu_j$ and defined $\xi_j$ to
be \eqref{eq:conditional-means} rather than \eqref{eq:xi}.  A referee objected
to defining $\xi_j$ to be \eqref{eq:conditional-means} because
\eqref{eq:conditional-means} does not define a parameter, being a function
of both data and parameters, and in statistics greek letters are supposed to
denote parameters.  Our definitions here belatedly admit the referee was right.}
\begin{subequations}
\begin{align}
   \mu_j & = E(y_j)
   \\
   \xi_j & = E(y_j \mid y_{p(j)} = 1)
   \label{eq:xi}
\end{align}
\end{subequations}
In words, $\mu_j$ is the unconditional mean value of $y_j$ and $\xi_j$ is
the mean value of one of the IID random variables of which $y_j$
is the sum.  We call the vector $\mu$ with components $\mu_j$ the
\emph{unconditional mean value parameter vector}
and the vector $\xi$ with components $\xi_j$
the \emph{conditional mean value parameter vector}.
Because the expectation of a sum is
the sum of the expectations,
\begin{equation} \label{eq:conditional-means}
   E(y_j \mid y_{p(j)}) = y_{p(j)} \xi_j,
\end{equation}
and by the iterated expectation theorem
\begin{equation} \label{eq:means}
   \mu_j = \mu_{p(j)} \xi_j.
\end{equation}
Equation \eqref{eq:means} is important.  It relates conditional
and unconditional mean values in aster models.  It is peculiar to models
having the ``predecessor is sample size'' property; it has no analog
for general statistical models.  Iterating it we obtain
\begin{equation} \label{eq:xi-to-mu}
\begin{split}
   \mu_j & = \xi_j \mu_{p(j)}
   \\
   & = \xi_j \xi_{p(j)} \mu_{p(p(j))}
   \\
   & = \xi_j \xi_{p(j)} \xi_{p(p(j))} \mu_{p(p(p(j)))}
\end{split}
\end{equation}
and so forth.  To find the unconditional mean value for $y_j$ we multiply
the conditional mean values for all the arrows (each representing a conditional
distribution) going back to the predecessor of predecessor of predecessor etc.\
that is a constant.  The expectation of a constant is that constant, so
when $k = p(p(p(j)))$ or whatever (perhaps more $p$'s) is such that $y_k$ is
a constant, then $\mu_k = y_k$, and this allows
calculation of all the $\mu_j$ as a function of all the $\xi_j$.

Conversely, solving \eqref{eq:means} for $\xi_j$ gives
\begin{equation} \label{eq:mu-to-xi}
   \xi_j = \frac{\mu_j}{\mu_{p(j)}}
\end{equation}
except in the case, which never occurs in practice,
where the denominator on the right-hand side is zero
(division by zero is undefined), and this allows
calculation of all the $\xi_j$ as a function of all the $\mu_j$.

Thus conditional mean values collectively determine unconditional mean values
and vice versa via \eqref{eq:xi-to-mu} and \eqref{eq:mu-to-xi}.
For example, for the graph \eqref{gra:one} we have
\begin{align*}
    \mu_1 & = \xi_1 \\
    \mu_2 & = \xi_2 \xi_1 \\
    \mu_3 & = \xi_3 \xi_2 \xi_1 \\
    \mu_4 & = \xi_4 \xi_3 \xi_2 \xi_1 \\
    \mu_5 & = \xi_5 \xi_1 \\
    \mu_6 & = \xi_6 \xi_2 \xi_1 \\
    \mu_7 & = \xi_7 \xi_3 \xi_2 \xi_1 \\
    \mu_8 & = \xi_8 \xi_4 \xi_3 \xi_2 \xi_1 \\
\end{align*}
and
\begin{align*}
    \xi_1 & = \mu_1 \\
    \xi_2 & = \mu_2 / \mu_1 \\
    \xi_3 & = \mu_3 / \mu_2 \\
    \xi_4 & = \mu_4 / \mu_3 \\
    \xi_5 & = \mu_5 / \mu_1 \\
    \xi_6 & = \mu_6 / \mu_2 \\
    \xi_7 & = \mu_7 / \mu_3 \\
    \xi_8 & = \mu_8 / \mu_4 \\
\end{align*}

We are not finished with the implications of ``predecessor as sample size''
but the rest will have to wait until Section~\ref{sec:expo},
when we will have developed more theory.

\subsection{Exponential Families of Distributions} \label{sec:expo-general}

The likelihood function of a family of distributions is just the probability
function thought of as a function of the parameters for one fixed data value
rather than as a function of data for one fixed parameter value.  Usually
its logarithm (log likelihood) is used.  For an aster model, the log likelihood
is the logarithm of \eqref{eq:factorize-too}
\begin{equation} \label{eq:aster-logl}
   l(\theta) = \sum_{j \in J} \log f_\theta(y_j \mid y_{p(j)}),
\end{equation}
where we have added explicit dependence on the parameter vector $\theta$
to the conditional probability functions.  There is one term in the
log likelihood for each arrow in the graph.

Because uses of the
log likelihood do not depend on constants, one may add or subtract a term
that does not depend on the parameters (but may depend on the data) without
effect on statistical inference.  This is usually done without comment;
additive terms in the log likelihood that do not contain parameters are
omitted.

A parametric family of distributions is said to be an \emph{exponential family}
if the log likelihood has the form
\begin{equation} \label{eq:logl}
   l(\theta) = \inner{y, \theta} - c(\theta),
\end{equation}
where $y$ is a vector statistic (a function of the data that does not depend
on parameters), $\theta$ is a vector parameter, $\inner{\fatdot, \fatdot}$ is
the bilinear form defined by \eqref{eq:bilinear}, and $c$ is a function of
the parameter vector called the \emph{cumulant function} of the family.

The vector statistic $y$ and vector parameter $\theta$ in \eqref{eq:logl}
need not be the originally given data and parameter; one may need to
do either a change-of-variable or change-of-parameter (or both) to get the
log likelihood into exponential family form \eqref{eq:logl}.  To express
that $y$ and $\theta$ are the special statistic and parameter vectors that
appear in \eqref{eq:logl}, they are called the \emph{canonical statistic}
and \emph{canonical parameter}.

Exponential families have a simple relationship with IID (which, recall from
Section~\ref{sec:intro},
means independent and identically distributed).\footnote{In aster models,
components of the response vector are neither independent nor identically
distributed (because of the graphical structure and because they have
different parameters), and in LM and GLM they are not identically distributed
(because they have different parameters).
Nevertheless, the property of exponential family models discussed in the
remainder of this section will be useful in constructing aster models
(Section~\ref{sec:expo} below).}
Suppose, using somewhat confusing notation for this paragraph only,
$z_1$, $\ldots$, $z_n$ are IID random vectors
(the confusion is that now subscripts denote different random vectors
rather than components of one vector)
all having the same exponential family distribution for which they are
the canonical statistic vectors and for which $\theta$ is
the canonical parameter vector.
Stochastic independence means
that the joint probability function is the product of the marginals
$$
   f_\theta(z_1, \ldots, z_n) = \prod_{i = 1}^n f_\theta(z_i)
$$
from which it follows from the log of a product being the sum of the logs
that the log likelihood is
$$
   l(\theta) = \sum_{i = 1}^n \log f_\theta(z_i)
$$
and because of the exponential family assumption each term of the log
likelihood is \eqref{eq:logl} with $y$ replaced by $z_i$ giving
\begin{equation} \label{eq:iid-expo}
\begin{split}
   l(\theta)
   & =
   \sum_{i = 1}^n \bigl[ \inner{z_i, \theta} - c(\theta) \bigr]
   \\
   & =
   \inner{z_1 + \cdots + z_n, \theta} - n c(\theta)
\end{split}
\end{equation}
Thus we again get an exponential family.
The canonical statistic vector for IID data is
the sum $z_1 + \cdots + z_n$ of the canonical statistic vectors $z_i$ for the
IID constituents and the canonical parameter vector $\theta$ is the same
for both.  Moreover, the cumulant function for the IID family is $n c(\theta)$,
just $n$ times the cumulant function for the IID constituents.
This relationship
is peculiar to exponential families; it has no analog for general statistical
models.

\subsection{Exponential Families and Aster Models} \label{sec:expo}

In an aster model, the conditional distribution associated with each arrow
in the graph and each term in \eqref{eq:aster-logl}
is required to be a one-parameter exponential family: $y_j$ has the
distribution of a sum of IID random variables in some exponential family,
the number of terms in the sum being $y_{p(j)}$ (this is the ``predecessor
is sample size'' property again).  From the relationship between IID and
exponential families in the previous section, this means the term of the
log likelihood associated with the $j$-th arrow is
\begin{equation} \label{eq:logl-one-term}
   y_j \theta_j - y_{p(j)} c_j(\theta_j)
\end{equation}
because we have a one-parameter family so the bilinear form becomes
just $y_j \theta_j$ and what was $n$ in \eqref{eq:iid-expo} is $y_{p(j)}$
here because of ``predecessor is sample size.''  Also we have subscripts on
the parameter and cumulant function because each arrow in the graph may
be associated with a different one-parameter exponential family, each
having a different parameter $\theta_j$ and different cumulant function $c_j$.

It is a trivial but not entirely obvious fact that \eqref{eq:logl-one-term}
works for the case where $y_{p(j)} = 0$ as well as for other cases.
When $y_{p(j)} = 0$ the assumption is that $y_j$ is the sum of zero IID random
variables, and (as mentioned at the beginning of Section~\ref{sec:sample-size})
a sum with zero terms is zero (by convention).  Thus the conditional
distribution of $y_j$ given $y_{p(j)} = 0$ is the distribution concentrated
at zero, that is, $y_j = 0$ with (conditional) probability one.
Since $\log(1) = 0$, the term in the log likelihood for the conditional
distribution of $y_j$ given $y_{p(j)}$ should be zero when $y_{p(j)} = 0$,
and that is what \eqref{eq:logl-one-term} gives, because
$y_{p(j)} = 0$ implies $y_j = 0$.
Thus, although $y_{p(j)} = 0$ is a special case, we do not need to deal with
it specially, since the general formula \eqref{eq:logl-one-term} works for
this special case too.\footnote{This observation explains why what many think
of as a missing data problem is not.  What is the number of offspring $y_j$ in
a time period when the individual is dead so $y_{p(j)} = 0$?  In
aster models $y_{p(j)} = 0$ implies $y_j = 0$ so $y_j$ is not missing but
known to be zero.  As we have just seen, the aster log
likelihood does the right thing in this case.\label{foot:zeros}}

Thus the log likelihood for the aster model is
\begin{equation} \label{eq:logl-cond}
   l(\theta)
   =
   \sum_{j \in J} \bigl[ y_j \theta_j - y_{p(j)} c_j(\theta_j) \bigr],
\end{equation}
and this works for all cases, including when some of the $y_j$ or $y_{p(j)}$
are zero.

Now we can explain the labels attached to the arrows in the graphs
\eqref{gra:one} and \eqref{gra:complicated}.  We call the labels
the \emph{annotation} of the graph.
Each denotes the distribution of one
of the IID random variables of which $y_j$ is the sum, that is,
it denotes the distribution
having cumulant function $c_j(\theta_j)$.
In \eqref{gra:one} and \eqref{gra:complicated}
``Ber'' indicates a Bernoulli distribution
and ``Poi'' indicates a Poisson distribution.
In \eqref{gra:complicated} ``0-Poi'' indicates a zero-truncated Poisson
distribution (Poisson conditioned on being nonzero).
It is perhaps somewhat
confusing that the labels do not denote the conditional distribution of $y_j$
given $y_{p(j)}$.  For example, if the distribution having cumulant function
$c_j(\theta_j)$ is Bernoulli, then the distribution having
cumulant function $y_{p(j)} c_j(\theta_j)$ is binomial with sample size
$y_{p(j)}$ and canonical parameter $\theta_j$.  It might seem more sensible
to name these distributions binomial rather than Bernoulli, since the name
binomial is more familiar to users, but this would not work in general.
For some distributions used in aster models, we have a name for
the distribution having cumulant function $c_j(\theta_j)$ but no name
for the distribution having cumulant function $y_{p(j)} c_j(\theta_j)$;
zero-truncated Poisson is an example.\footnote{Not only do we have no name
for the distribution of the sum of IID zero-truncated Poisson random variables,
we have no explicit formula for its probability function, but we do not need
one because of the relationship between IID and exponential families discussed
in Section~\ref{sec:expo-general}.  We know the cumulant function will be
$y_{p(j)} c_j(\theta_j)$, and that's all we need to know.  This ultimately
derives from the property that additive terms not involving the parameters can
be dropped from log likelihoods.  Here we are dropping terms that we do not
know how to calculate but do know do not contain the parameters.}

Since each arrow in the graph is allowed to have a different exponential
family distribution, some may be discrete distributions (Bernoulli, Poisson,
zero-truncated Poisson) and some may be continuous (normal).  However,
$y_j$ that are predecessor variables must be nonnegative-integer-valued
because they are sample sizes for their successors.\footnote{\citet{gws}
note that if the successor is infinitely divisible, for example, Poisson
or normal, then the predecessor can be positive-real-valued, but this
generalization has not been used in aster models
and may be ignored.\label{foot:infinitely-divisible}}
Thus only terminal nodes of the graph (those having no successors)
can be associated with continuous distributions.

\subsection{Reparameterization} \label{sec:geyer-great-idea}

Now we come to what is the only completely new idea in \citet{gws}.
The log likelihood \eqref{eq:logl-cond} does not have exponential family form
\eqref{eq:logl}.  Each term has this form, but their sum does not, because of
both $y_j$ and $y_{p(j)}$ being random in some terms.
Observe that
\eqref{eq:logl-cond} is linear in the components of the data
vector $y_j$ and $y_{p(j)}$ so the joint distribution
is an exponential family with $y_J$ as its canonical statistic.  To identify
the canonical parameters we rewrite the log likelihood collecting terms that
multiply the same component of $y$
\begin{equation} \label{eq:rewrite}
   l(\theta)
   =
   \sum_{j \in J} y_j \left[
   \theta_j - \sum_{\substack{m \in J \\ j = p(m)}} c_m(\theta_m) \right]
   -
   \sum_{\substack{m \in J \\ p(m) \in F}} y_{p(m)} c_m(\theta_m).
\end{equation}
Recall that the $y_j$ are random when $j \in J$ and constant when $j \in F$
so that in \eqref{eq:rewrite} the random variables are the $y_j$ and the
constants are the $y_{p(m)}$.  Thus the canonical parameters for the
joint exponential family distribution of the data are
\begin{equation} \label{eq:theta-to-phi}
   \varphi_j = \theta_j - \sum_{\substack{m \in J \\ j = p(m)}} c_m(\theta_m),
   \qquad j \in J,
\end{equation}
which are the terms in square brackets in \eqref{eq:rewrite} that are the
coefficients of the $y_j$.  \citet{gws} observe that \eqref{eq:theta-to-phi}
is an invertible change-of-parameter, because \eqref{eq:theta-to-phi} can
be solved for $\theta_j$ giving
\begin{equation} \label{eq:phi-to-theta}
   \theta_j = \varphi_j + \sum_{\substack{m \in J \\ j = p(m)}} c_m(\theta_m),
\end{equation}
which expresses $\theta_j$ as a function of $\varphi_j$ and $\theta_m$ such
that $j = p(m)$, that is, components of $\theta$ for successors of $\theta_j$.
At \emph{terminal}
nodes of the graph (those having no successors) we have $\theta_j = \varphi_j$.
Now we can solve for $\theta_j$ whose successors are terminal nodes.
After that we can solve for $\theta_j$ whose successors or successors of
successors are terminal nodes, and so forth.
In fact, we can solve in any order
that finds components of $\theta$ for successors before predecessors,
whatever is computationally
convenient.  Thus \eqref{eq:phi-to-theta}, somewhat implicitly, defines
the vector $\theta$ having components $\theta_j$ as a function of
the vector $\varphi$ having components $\varphi_j$.

The cumulant function of the joint exponential family with canonical
parameter vector $\varphi$ must be given by
\begin{equation} \label{eq:cum-fun-aster}
   c(\varphi) =
   \sum_{\substack{m \in J \\ p(m) \in F}} y_{p(m)} c_m(\theta_m),
\end{equation}
the right-hand side being the leftovers, terms in \eqref{eq:rewrite} that
have not been used in \eqref{eq:theta-to-phi}.
As the subscripts on the summation sign
say, all of the $y_{p(m)}$ appearing in this cumulant function have
$p(m) \in F$, hence these denote constants not random variables so
\eqref{eq:cum-fun-aster} does define a non-random function.

Using the fact that the cumulant function for the Bernoulli family is
$$
   c(\theta) = \log(1 + e^\theta)
$$
and the cumulant function for the Poisson family is
$$
   c(\theta) = e^\theta
$$
we get for the graph \eqref{gra:one}
\begin{equation} \label{eq:big-theta-to-phi}
\begin{split}
   \varphi_8 & = \theta_8 \\
   \varphi_7 & = \theta_7 \\
   \varphi_6 & = \theta_6 \\
   \varphi_5 & = \theta_5 \\
   \varphi_4 & = \theta_4 - e^{\theta_8} \\
   \varphi_3 & = \theta_3 - e^{\theta_7} - \log(1 + e^{\theta_4}) \\
   \varphi_2 & = \theta_2 - e^{\theta_6} - \log(1 + e^{\theta_3}) \\
   \varphi_1 & = \theta_1 - e^{\theta_5} - \log(1 + e^{\theta_2})
\end{split}
\end{equation}
and
\begin{equation} \label{eq:big-phi-to-theta}
\begin{split}
   \theta_8 & = \varphi_8 \\
   \theta_7 & = \varphi_7 \\
   \theta_6 & = \varphi_6 \\
   \theta_5 & = \varphi_5 \\
   \theta_4 & = \varphi_4 + e^{\theta_8} \\
            & = \varphi_4 + e^{\varphi_8} \\
   \theta_3 & = \varphi_3 + e^{\theta_7} + \log(1 + e^{\theta_4}) \\
            & = \varphi_3 + e^{\varphi_7}
                + \log(1 + \exp(\varphi_4 + e^{\varphi_8})) \\
   \theta_2 & = \varphi_2 + e^{\theta_6} + \log(1 + e^{\theta_3}) \\
            & = \varphi_2 + e^{\varphi_6}
                + \log(1 + \exp(
            \varphi_3 + e^{\varphi_7}
                + \log(1 + \exp(\varphi_4 + e^{\varphi_8})))) \\
   \theta_1 & = \varphi_1 + e^{\theta_5} + \log(1 + e^{\theta_2}) \\
            & = \varphi_1 + e^{\varphi_5} + \log(1 + \exp(
            \varphi_2 + e^{\varphi_6}
                + \log(1 + \exp(
            \varphi_3 + e^{\varphi_7}
            \\
            & \hphantom{=} \qquad
                + \log(1 + \exp(\varphi_4 + e^{\varphi_8}))))))
\end{split}
\end{equation}
and
\begin{align*}
   c(\varphi)
   & =
   \log(1 + e^{\theta_1})
   \\
   & =
   \log(1 + \exp(
            \varphi_1 + e^{\varphi_5} + \log(1 + \exp(
            \varphi_2 + e^{\varphi_6}
            \\
            & \hphantom{=} \qquad
                + \log(1 + \exp(
            \varphi_3 + e^{\varphi_7}
                + \log(1 + \exp(\varphi_4 + e^{\varphi_8}))))))
   ))
\end{align*}
As one can see, this reparameterization is quite complicated
and even more so for
larger graphs; if we were to repeat the calculations above for the graph
\eqref{gra:complicated} the formula for the cumulant function would not
fit on one page.  However, this reparameterization is very systematic
and easily handled by the
computer program that does aster models
\citep{package}.  Users do not need to know these formulas; the computer
deals with them.  We show them only for readers who want a concrete example.

Now we have put the log likelihood of an aster model in the exponential
family form
\begin{equation} \label{eq:logl-varphi}
   l(\varphi) = \inner{y, \varphi} - c(\varphi),
\end{equation}
where the cumulant function is given by \eqref{eq:cum-fun-aster} and
we have written the canonical statistic vector as $y$ rather than $y_J$
and the canonical parameter vector as $\varphi$ rather than $\varphi_J$,
which we will do from now on.
We see that \eqref{eq:logl-varphi}
is just like the log likelihood for a general exponential
family \eqref{eq:logl} except that the parameter has changed
from $\theta$ to $\varphi$.

O.~K\@.  We have found that the joint distribution of an aster model is an
exponential family and found the canonical parameters and cumulant function
of the joint distribution.  What was the point of that?

\subsection{Means, Variances, Convexity, and Concavity} \label{sec:convex}

The cumulant function of an exponential family determines means and variances
of the canonical statistic
\begin{subequations}
\begin{align}
   E_\varphi(y) & = \nabla c(\varphi)
   \label{eq:cum-one}
   \\
   \var_\varphi(y) & = \nabla^2 c(\varphi)
   \label{eq:cum-two}
\end{align}
\end{subequations}
where $\mu = E_\varphi(y)$ denotes the vector
whose components are $\mu_j = E_\varphi(y_j)$ and $\var_\varphi(y)$ denotes
the matrix whose components\footnote{The \emph{variance} of a random variable
$y_j$ with mean $\mu_j$ is
\begin{equation} \label{eq:variance}
   \var_\varphi(y_j) = E_\varphi\{ (y_j - \mu_j)^2 \}.
\end{equation}
The \emph{covariance} of random variables $y_j$ and $y_k$ with means $\mu_j$
and $\mu_k$ is \eqref{eq:covariance}.  Note that when $j = k$
\eqref{eq:covariance} reduces to \eqref{eq:variance}, so the covariance of
a random variable with itself is the variance.  For this reason there is no
generally accepted terminology for the matrix with components
\eqref{eq:covariance}.  We call it the ``variance matrix'' because it plays
the same role for a random vector $y$ as variance does for a random variable.
But other people call it the ``covariance matrix'' or the ``variance-covariance
matrix'' or (annoyed at the confusion, inventing a new name) the ``dispersion
matrix.''}
are
\begin{equation} \label{eq:covariance}
   \cov_\varphi(y_j, y_k) = E_\varphi\{ (y_j - \mu_j) (y_k - \mu_k) \}
\end{equation}
and where $\nabla c(\varphi)$ denotes the vector
whose components are $\mu_j = \partial c(\varphi) / \partial \varphi_j$
and $\nabla^2 c(\varphi)$ denotes the matrix whose components
are
$$
   \partial^2 c(\varphi) / \partial \varphi_j \partial \varphi_k
$$
\citep[Theorem~8.1]{barndorff}.  The identities \eqref{eq:cum-one}
and \eqref{eq:cum-two} have a number of important consequences.  First
(at least for aster models where the cumulant functions have known expressions),
they allow probabilistic calculations about means and variances, which may
be difficult to compute (involving sums and integrals) with derivatives,
which are easier to compute.
Second, they have a number of important theoretical consequences.

A variance matrix is positive semi-definite and positive definite unless
the distribution of the random vector is concentrated on
a hyperplane, in which case some components can be expressed as linear
functions of other components and eliminated.  Thus we can always arrange
that the distribution of the canonical statistic vector $y$ is non-degenerate
(not concentrated on a hyperplane), in which case the variance matrix in
\eqref{eq:cum-two} is positive definite, which in turn implies that the
cumulant function is strictly convex \citep[Theorem~2.14]{raw}.

Consider the map $h$ between canonical parameter values $\varphi$ and mean
values $\mu$ defined by
$$
   h(\varphi) = E_\varphi(y) = \nabla c(\varphi).
$$
Fix a possible value $\mu^*$ of the mean vector $\mu = h(\varphi)$,
and define the function
\begin{equation} \label{eq:qfoo-zero}
   q(\varphi) = \inner{\mu^*, \varphi} - c(\varphi)
\end{equation}
which is just the log likelihood \eqref{eq:logl-varphi} with $y$ replaced by
$\mu^*$.  Applying \eqref{eq:cum-one} and \eqref{eq:cum-two} gives
\begin{equation} \label{eq:qfoo-one}
\begin{split}
   \nabla q(\varphi)
   & =
   \mu^* - \nabla c(\varphi)
   \\
   & =
   \mu^* - h(\varphi)
\end{split}
\end{equation}
and
\begin{equation} \label{eq:qfoo-two}
   \nabla^2 q(\varphi) = - \nabla^2 c(\varphi).
\end{equation}
Equation \eqref{eq:qfoo-two} implies $q$ is
a strictly concave function,\footnote{A function $f$ is concave if $- f$ is
convex, and strictly concave if $- f$ is strictly convex.  Here $q$ is
strictly concave because $- q$ is strictly convex,
because $\nabla^2 c(\varphi)$ is positive definite.}
which in turn implies the maximizer $\varphi^*$ of $q$ is unique if it
exists \citep[Theorem~2.6]{raw}.  From calculus we know that the first
derivative is zero at the maximum, so by \eqref{eq:qfoo-one} $\varphi^*$
is a solution of
$$
   \mu^* = h(\varphi).
$$
By assumption $\mu^*$ is a possible mean value, hence there exists
a $\varphi^*$ such that $\mu^* = h(\varphi^*)$ and the concavity argument
shows that there is only one such $\varphi^*$.  This proves $h$ is an
invertible mapping.  Each mean value $\mu$ corresponds to a unique
canonical parameter value $\varphi$.

The same argument applied to the one-parameter exponential families
corresponding to arrows in the graphical model shows
that the mapping between $\theta_j$ and $\xi_j$ given by
$$
   \xi_j = E_{\theta_j}(y_j \mid y_{p(j)} = 1) = c_j'(\theta_j),
$$
where the prime denotes
differentiation $c_j'(\theta_j) = d c_j(\theta j) / d \theta_j$,
is also an invertible mapping.  Each conditional mean value
$\xi_j$ corresponds to a unique canonical parameter value $\theta_j$ of
the corresponding one-parameter conditional exponential family.

\subsection{Many Parameterizations}

An aster model has four parameterizations that we have introduced so far:
conditional and unconditional mean value parameter vectors $\xi$ and $\mu$
and what \citet{gws} call conditional and unconditional canonical parameter
vectors $\theta$ and $\varphi$.

In Section~\ref{sec:sample-size} we established that the change of parameter
$\xi \to \mu$ is invertible.
In Section~\ref{sec:geyer-great-idea} we established that
the change of parameter $\theta \to \varphi$ is invertible.
At the end of the preceding section we established that
the change of parameter $\varphi \to \mu$ is invertible
and that the change of parameter $\theta \to \xi$ is invertible.

Thus all four parameterizations are
equally valid.  Each is interconvertible with the others.
The changes-of-parameter $\xi \to \mu$ and $\mu \to \xi$ are given by
explicit formulas \eqref{eq:xi-to-mu} and \eqref{eq:mu-to-xi}.
The changes-of-parameter $\theta \to \varphi$ and $\varphi \to \theta$
are given by
explicit formulas \eqref{eq:theta-to-phi} and \eqref{eq:phi-to-theta}.
The changes-of-parameter $\theta \to \xi$ and $\varphi \to \mu$
are given by explicit formulas
\begin{align*}
   \xi_j & = c_j'(\theta_j)
   \\
   \mu & = \nabla c(\varphi)
\end{align*}

The changes-of-parameter $\xi \to \theta$ and $\mu \to \varphi$ are different.
We know that each possible unconditional mean value parameter vector $\mu^*$
corresponds to a unique unconditional canonical parameter vector $\varphi^*$,
but we can only find that $\varphi^*$ by maximizing the function $q$ defined
in \eqref{eq:qfoo-zero}.  This optimization problem may not have a solution
given by a formula as a function of $\mu^*$.  We know the function mapping
$\mu \to \varphi$ exists.  We even know it is infinitely differentiable by
the inverse function theorem of real analysis.  But we know nothing else about
it other than that we can calculate its value at any point $\mu^*$ by running
a computer optimization algorithm to find the $\varphi^*$ where $q$ achieves
its maximum.

Similarly, in general, we can only find the unique $\theta_j^*$
such that $\xi_j^*$ is the corresponding conditional mean value parameter for
the $j$-th arrow in the graph by running a computer optimization algorithm
to find the $\theta_j^*$ where the function
$$
   q_j(\theta_j) = \xi_j^* \theta_j - c_j(\theta_j)
$$
achieves its maximum.

Most but not all of the one-parameter families
implemented in the current version of the aster model do have expressions
as formulas for the mapping $\xi_j \to \theta_j$.
For example, for Bernoulli arrows
\begin{align*}
   \xi_j & = \frac{e^{\theta_j}}{1 + e^{\theta_j}}
   \\
   \theta_j & = \logit(\xi_j)
\end{align*}
and for Poisson arrows
\begin{align*}
   \xi_j & = e^{\theta_j}
   \\
   \theta_j & = \log(\xi_j)
\end{align*}
Thus for a graph like \eqref{gra:one} having only such arrows we do have (very
complicated) expressions as formulas for the map $\mu \to \varphi$ going
by the route $\mu \to \xi \to \theta \to \varphi$.

In contrast, for the zero-truncated Poisson distribution
$$
   \xi_j = \frac{\exp(\theta_j)}{1 - \exp(- \exp(\theta_j))}
$$
and this cannot be solved for $\theta_j$ to give a formula expressing
$\theta_j$ as a function of $\xi_j$.  Thus for a graph like
\eqref{gra:complicated} having zero-truncated Poisson arrows, there is
no formula expressing the mapping $\xi \to \theta$
or the mapping $\mu \to \varphi$.

It must be admitted that the names ``conditional canonical parameter''
and ``unconditional canonical parameter'' for the vectors $\theta$ and $\varphi$
are somewhat misleading because the two parameters are not as analogous
as the parallel terminology makes them seem.  Pedantically, $\theta$ is
the vector whose components are the canonical parameters of the one-parameter
exponential families associated with the arrows of the graph, whereas
$\varphi$ is the canonical parameter vector for the joint exponential family
distribution of the aster model.
Similarly, the names ``conditional mean value parameter''
and ``unconditional mean value parameter'' for the vectors $\xi$ and $\mu$
are somewhat misleading because the two parameters are not as analogous
as the parallel terminology makes them seem.  Pedantically, $\xi$ is
the vector whose components $\xi_j$ are the mean value parameters for sample
size one of the one-parameter
exponential families associated with the arrows
of the graph,\footnote{Recall that \label{foot:xi}
\begin{align*}
    \xi_i & = E(y_j \mid y_{p(j)} = 1)
    \\
    y_{p(j)} \xi_i & = E(y_j \mid y_{p(j)})
\end{align*}}
whereas
$\mu$ is the mean value parameter vector for the joint exponential family
distribution of the aster model.

\subsection{Canonical Linear Submodels and Maximum Likelihood}
\label{sec:submodel}

It may be disconcerting to the reader to be told the models described
up to this point are uninteresting in themselves because they have too
many parameters (one parameter per component of the response vector,
no matter which parameterization $\varphi$, $\theta$, $\mu$, or $\xi$
is used) to estimate well.  As is the case with LM and GLM, only submodels
with fewer parameters are interesting.
We call the models described up to this point \emph{saturated}.
We are interested in \emph{canonical linear submodels} having parameter
vector $\beta$ related to the other parameters by
\begin{equation} \label{eq:canonical-linear-submodel}
   \varphi = M \beta,
\end{equation}
which is just like \eqref{eq:glm} except that in GLM there is no requirement
that the linear predictor vector be the canonical parameter vector, and here
we do require this.

Exponential families have a simple relationship
with canonical linear submodels.
The log likelihood for the new parameter vector $\beta$
is \eqref{eq:logl-varphi}
with $M \beta$ plugged in for $\varphi$ giving
\begin{equation} \label{eq:logl-beta-foo}
   l(\beta) = \inner{y, M \beta} - c(M \beta).
\end{equation}
Observe that
$$
   \inner{y, M \beta} = \sum_{j \in J} y_j \sum_{k = 1}^p m_{j k} \beta_k
$$
where $p$ is the dimension of $\beta$ and $m_{j k}$ are the components of
the matrix $M$.  If we reverse the order of summation, thinking of the sum
over $k$ as the one for an inner product and the one over $j$ as the one
for a matrix-vector multiplication we get
$$
   \inner{M^T y, \beta} = \sum_{k = 1}^p \beta_k \sum_{j \in J} m_{j k} y_j
$$
where $M^T$ denotes the transpose of $M$
(the matrix with $j, k$ element $m_{k j}$).

This algebraically trivial transformation has very important statistical
consequences.  When we rewrite \eqref{eq:logl-beta-foo} as
\begin{equation} \label{eq:logl-beta}
   l(\beta) = \inner{M^T y, \beta} - c(M \beta)
\end{equation}
we see we have exponential family form.  The submodel of the saturated model
determined by the change-of-parameter $\varphi = M \beta$ is itself an
exponential family with canonical statistic vector $M^T y$, canonical
parameter vector $\beta$,
and cumulant function $c_{\text{submodel}}(\beta) = c(M \beta)$.

Thus the log likelihood \eqref{eq:logl-beta} is a strictly concave function
and the MLE $\hat{\beta}$, the maximizer of \eqref{eq:logl-beta} is unique
if it exists (from exponential family theory developed
in Section~\ref{sec:convex}).\footnote{The MLE need not exist.  Conditions
for when it does not were first described by \citet{barndorff} and generalized
in the author's unpublished thesis \citep{thesis}.  They involve a fascinating
interplay of the geometry of convex polyhedra, the geometry of exponential
family log likelihood functions, linear programming, and general statistical
theory.  A recent paper \citep{gdor} gives a complete computational solution
for when the MLE does not exist and describes what to do when it does not.
That paper is only about GLM, but does point the way toward corresponding
theory for aster models.\label{foot:barndorff-complete}}
The principle of invariance of maximum likelihood gives MLE
of all other parameters.\footnote{This principle says that if $\theta$ and
$\psi$ are two parameters related by an invertible change-of-parameter,
say $\psi = g(\theta)$, then the MLE are related
by $\hat{\psi} = g(\hat{\theta})$.  Thus in the case under discussion
$\hat{\varphi} = M \hat{\beta}$, and all other parameters follow, for example,
$\hat{\mu} = \nabla c(\hat{\varphi})$.\label{foot:invariance}}

The log likelihood is an extremely complicated function of $\beta$, and
there is no formula expressing $\hat{\beta}$ as a function of $y$.
We know nothing about the MLE $\hat{\beta}$
other than that we can calculate its value for any data $y$ by running
a computer optimization algorithm to find the $\hat{\beta}$ where $l$ achieves
its maximum.

We do know something more about the MLE $\hat{\mu}$ of the mean value
parameter.  Differentiating \eqref{eq:logl-beta} gives
\begin{align*}
   \nabla l(\beta)
   & =
   M^T y - M^T \nabla c(M \beta)
   \\
   & =
   M^T y - M^T h(M \beta)
\end{align*}
(the $M^T$ in the second term coming from the chain rule for differentiation),
where $h$ is the map $\mu = h(\varphi) = \nabla c(\varphi)$ introduced in
Section~\ref{sec:convex}.
The derivative of the log likelihood is zero at the maximum, hence the MLE
satisfies
$$
   M^T y = M^T h(M \hat{\beta})
$$
or $M^T y = M^T \hat{\mu}$.  This is a general property of exponential
families not peculiar to aster models: the MLE of
the \emph{mean value parameter} satisfies a property
$$
   \text{observed}
   =
   \text{expected}
$$
which is shorthand for observed value of the canonical statistic vector
(for the submodel, here $M^T y$) equals expected value of the the canonical
statistic vector
with the MLE parameter value plugged in (here $M^T \hat{\mu}$).

``Observed = expected'' is not useful for calculation,
since the matrix $M$ is not full rank, so if we simply seek solutions to
the linear system of equations $M^T y = M^T \mu$, there are many solutions,
only one of which is the MLE $\hat{\mu}$.  The only way to find $\hat{\mu}$
is via the path already described, first find $\hat{\beta}$ by maximizing
the log likelihood \eqref{eq:logl-beta} and then use the transformation
$\hat{\mu} = h(M \hat{\beta})$.

Thus ``observed = expected'' is useful only for interpretation.  It is
the only simple relationship between data $y$ and the MLE $\hat{\mu}$ of
some parameter vector.  Note that we have to be careful in using
``observed = expected'' because it applies only to the canonical statistic
vector of \emph{the model being used}, which in this case is not the
saturated model with canonical statistic vector $y$ and canonical parameter
vector $\varphi$ but rather the canonical linear submodel with canonical
statistic vector $M^T y$ and (submodel) canonical parameter vector $\beta$.
So in this case ``observed = expected'' says the observed value of the
submodel canonical statistic vector $M^T y$ is equal to its expected value
$M^T \mu$ with the MLE $\hat{\mu}$ plugged in for the parameter vector $\mu$.

% There is a beautiful mathematical duality between canonical statistics
% and canonical parameters of exponential families explained in Chapter~1
% of \citet{thesis} and this same duality holds between expectations of
% canonical statistics, which are mean value parameters $\mu = E_\varphi(y)$,
% and canonical parameters $\varphi$.

We now have added a fifth parameter vector $\beta$ to go with the four
$\theta$, $\varphi$, $\xi$, and $\mu$ we already had.  This
plethora of parameterizations is unique to aster models.  Because there is
no difference between conditional and unconditional in GLM (components
of the response vector being independent), conditional and unconditional
are the same so $\theta = \varphi$ and $\xi = \mu$.  This
leaves only three parameter vectors $\beta$, $\varphi$, and $\mu$
for GLM.\footnote{In GLM the linear predictor vector $\eta$, need not be
the canonical parameter vector $\varphi$, but there are still only three
parameter vectors $\beta$, $\eta$, and $\mu$.}
Because for the normal distribution canonical
and mean value parameters are the same $\varphi = \mu$,
this leaves only two parameter vectors $\beta$ and $\mu$ for LM.

Multiple parameterizations are considered confusing to students in applied
courses about GLM and LM, so they are disguised
by changing the language.  The components of $\beta$ are called parameters,
but $\varphi$ is called the linear predictor vector and $\mu$ the mean
vector, and that they are also parameters is not mentioned.
Estimates $\hat{\varphi}$ and $\hat{\mu}$ are called
``predicted values'' (although there is no sense of future
in this ``prediction'' and they are, in fact, MLE of the corresponding
parameters).  This is very anti-theoretical.  It clouds
the distinction between estimates $\hat{\mu}$ and parameters $\mu$, which
is the fundamental issue in statistics.\footnote{In teaching undergraduate
introductory statistics, I try to say ``the sample is not the population''
at least once a day after these terms are introduced and also say
``estimates are not parameters''
at least once a day after these terms are introduced.  This is related to
the three fundamental errors of statistics.
\begin{enumerate}
\item The worst error is to not use data (anecdotes are not evidence).
\item The next worst error is to confuse sample and population or estimates
and parameters: if a poll says Jones is ahead of Smith by 51 to 49 percent,
then the erroneous conclusion is to say it's all over and the election might
as well not be held, even though the margin of error of the poll is 3 percent.
\item The next worst error (and a very subtle one) is to fail to interpret
statistical confidence intervals correctly.  The slogan here is ``it's called
a 95\% confidence interval because it misses 5\% of the time.''  If a poll with
3\% margin of error says Jones is ahead of Smith by 54\% to 46\%,
then the erroneous conclusion is to say it's all over and the election might
as well not be held.  The true unknown population percentages could be
Jones 49\% and Smith 51\% even though the poll says Jones 54\% and Smith 46\%
and the margin of error is 3\%.  In fact the probability that the poll results
differ from the (unknown) truth by more than the margin of error is 5\%
(that's the subtle point).
\end{enumerate}
All of introductory statistics in one footnote!\label{foot:intro}}

Our view, in teaching aster models, is that this attempt at avoiding confusion
is itself a source of confusion.  In teaching aster models we say that all
five parameter vectors $\beta$, $\theta$, $\varphi$, $\xi$, and $\mu$ are
parameters.  That is, they are vectors that specify particular probability
models within
families of probability models.  Each can be converted to any of the others
by the changes-of-parameter discussed above.
Tests of statistical hypotheses or confidence intervals can be performed
for any of these parameters.
If one doesn't call $\mu_j$ a parameter, then what does one call confidence
intervals of the form $\hat{\mu}_j \pm \text{margin of error}$?
It is fundamental in statistics that confidence intervals are (interval)
estimates of parameters, thus the thing being estimated $\mu_j$ must be
a parameter.  Avoiding calling $\mu$ a parameter vector can only lead
to confusion about what confidence intervals are.

A fine point about the many parameterizations should be mentioned.
For a saturated aster model there are four parameterizations
$\theta$, $\varphi$, $\mu$, and $\xi$.  All have the same dimension, the
dimension $n$ of the response vector.  For a canonical linear submodel,
the parameter $\beta$ has dimension $p < n$.  The other parameterizations
must also have dimension $p$.  The set of allowed values of $\varphi = M \beta$
is a $p$-dimensional vector subspace
of $n$-dimensional Euclidean space.
Since the maps $\varphi \to \mu$, $\varphi \to \theta$, and $\theta \to \xi$
are nonlinear but smooth, the sets of allowed values of $\mu$, $\theta$, and
$\xi$ are different $p$-dimensional manifolds in $n$-dimensional Euclidean
space.\footnote{This is true of most aster models.
The canonical parameter space of most one-parameter exponential families used
for aster models is the whole real line $\real$, but may be an open interval.
The only example of the latter currently implemented
is the negative binomial and truncated negative binomial
families which have $- \infty < \theta_j < 0$.

Consider first the implications for \emph{saturated aster models}.
The allowed values of the conditional canonical parameter vector $\theta$
is usually all of $n$-dimensional Euclidean space $\real^n$ but may be only
an open subset of $\real^n$.  The mappings $\theta \to \varphi$,
$\varphi \to \mu$, and $\theta \to \xi$ and their inverse mappings are all
smooth and hence map open sets to open sets.  It can be seen from \eqref{eq:theta-to-phi} that when the set of allowed values of $\theta$ is all of $\real^n$,
then so is the set of allowed values of $\varphi$.
Typically, the sets of allowed values of the mean value parameter vectors
$\mu$ and $\xi$ are not all of $\real^n$ but only open subsets thereof.

Now consider the implications for \emph{canonical linear submodels}.
If the set of allowed values of $\varphi$ for the saturated model is all
of $\real^n$, the set of allowed $\beta$ values is all of $\real^p$,
and the set of allowed values of $\varphi$ for the submodel is a
$p$-dimensional vector subspace of $\real^n$.
If the set of allowed values of $\varphi$ for the saturated model is not all
of $\real^n$ but only an open subset thereof,
the set of allowed values of $\beta$
is the open subset of $\real^p$ consisting of $\beta$ such that
$\varphi = M \beta$ is allowed.
The set of allowed values of $\varphi$ is a flat $p$-dimensional submanifold
(an open subset of a $p$-dimensional vector subspace of $\real^n$).
The sets of allowed values of $\theta$, $\mu$, and $\xi$ are each curved
$p$-dimensional submanifolds of $\real^n$, and not much more can be said about
them.

\label{foot:restrict}
One of the motivations for modeling canonical parameters linearly, discussed
at the end of the introduction, was to avoid restrictions, but we have seen
that some aster models (those with negative binomial arrows) have restrictions.
What does that do to the method of maximum likelihood?  Not much.
An exponential family is said to be \emph{regular} if the set of allowed values
of the canonical parameter is open.  All exponential families used with aster
models are regular.  For a regular exponential family, the log likelihood goes
to minus infinity as $\varphi$ goes to the boundary
of its set of allowed values, and this implies the MLE for $\beta$ is still
a solution of $\nabla l(\beta) = 0$, and the MLE for $\mu$ still satisfies the
``observed = expected'' property $M^T y = M^T \hat{\mu}$
\citep[Corollary~9.6]{barndorff}.
Optimization software for finding $\hat{\beta}$ need not handle explicit
constraints; it need only avoid steps outside
the set of allowed $\beta$ values by shortening such steps until they don't
go outside.}

\subsection{More on Monotonicity} \label{sec:moron}

\citet{gws} also consider what they call \emph{conditional aster models}
defined by $\theta = M \beta$ as opposed to \emph{unconditional aster models}
defined by $\varphi = M \beta$.  Conditional aster models do not
behave as well statistically as unconditional aster models and have been little
used.  \citet{aster2} did provide one conditional aster model example,
but they could have used an unconditional aster model instead.
Their main reason for choosing a conditional aster model
was just to provide an example of one.\footnote{This was not clearly stated
in the paper because of the convention of
scientific writing that every procedure be described as the unquestionably
correct way to go even though in truth many ways are reasonable.  Only by
looking at papers by many different authors does one see that there are
multiple reasonable ways to proceed in most situations.
The first line of \emph{Tao Te Ching}, literally translated as
``tao that can tao is not constant tao,'' can be more loosely translated as
``methodology that is useful cannot be unchanging.''  Science follows tao,
but conventional scientific writing is about as untaoistic as it is possible
to be.}

The choice between conditional and unconditional aster models is
dictated by which kind of mean value parameter is more important.
As discussed at the end of Section~\ref{sec:intro}, there is a multivariate
monotone relationship between the unconditional canonical and mean value
parameter vectors
of an aster model.  If $\varphi_1$ and $\varphi_2$ are two distinct possible
values of the unconditional canonical parameter vector and $\mu_1$ and $\mu_2$
are the corresponding values of the unconditional mean value parameter vector,
then
$$
   \inner{\mu_1 - \mu_2, \varphi_1 - \varphi_2} > 0.
$$
Of course, the same holds true for the one-parameter conditional exponential
families associated with each arrow of the graph.  If $\theta_1$ and $\theta_2$
are two distinct possible values of the conditional canonical parameter vector
and $\xi_1$ and $\xi_2$ are the
corresponding values of the conditional mean value parameter vector, then
$$
   \inner{\xi_1 - \xi_2, \theta_1 - \theta_2} > 0,
$$
in fact, this holds componentwise
$$
   (\xi_{1 j} - \xi_{2 j}) (\theta_{1 j} - \theta_{2 j}) > 0,
   \qquad j \in J,
$$
where the subscripts $j$ denote components.  So the question is: which
multivariate monotone relationship does one want?  The choices are
\begin{subequations}
\begin{equation} \label{eq:multi-mono-mu}
   \inner{\mu_1 - \mu_2, M \beta_1 - M \beta_2} > 0
\end{equation}
for unconditional aster models or
\begin{equation} \label{eq:multi-mono-xi}
   \inner{\xi_1 - \xi_2, M \beta_1 - M \beta_2} > 0
\end{equation}
\end{subequations}
for conditional aster models.  In the vast majority of applications
one wants \eqref{eq:multi-mono-mu}.

As a concrete example, consider the aster model with graph \eqref{gra:one}
and suppose the primary question of scientific interest is the relationship
between Darwinian fitness, defined to be the total number of offspring produced,
$y_5 + y_6 + y_7 + y_8$ in \eqref{gra:one}, and a quantitative covariate $x$,
which is some phenotypic characteristic of the organism.\footnote{Pedantically,
one should say that $y_5 + y_6 + y_7 + y_8$ is only the best surrogate of
Darwinian fitness available in this experiment or observational study,
but we will not bother with that.}
Also we should distinguish between observed fitness $y_5 + y_6 + y_7 + y_8$
and expected fitness $\mu_5 + \mu_6 + \mu_7 + \mu_8$.
The primary question of scientific interest is more precisely stated
as what is expected fitness expressed as a function of $x$?  This function
is called the \emph{fitness landscape} or \emph{adaptive landscape}.

We have been a bit sloppy in writing expected fitness as
$\mu_5 + \mu_6 + \mu_7 + \mu_8$ forgetting that \eqref{gra:one} is only
the graph for one individual.  Every other individual has an isomorphic
graph (same shape but different numbers on the variables).  To be precise
about this, we introduce two discrete variables $u_j$ and $v_j$ which describe
where in the graph a variable $y_j$ is.  First, $v_j$ has values 1, $\ldots$, 8
and says which node in \eqref{gra:one} $y_j$ corresponds to.
Second, $u_j$ has values 0 or 1 and says which kind of node in \eqref{gra:one}
$y_j$ corresponds to: $u_j = 0$ for survival nodes ($v_j = 1$, 2, 3, 4)
and $u_j = 1$ for offspring count nodes ($v_j = 5$, 6, 7, 8).
Now we write our canonical linear model equations
concretely as
\begin{equation} \label{eq:modmat}
   \varphi_j = \beta_{v_j} + u_j (\beta_9 x_j + \beta_{10} x_j^2)
\end{equation}
(abstractly, they are $\varphi = M \beta$),
where $x_j$ is the quantitative covariate value for the individual corresponding
to the maximal connected component containing $y_j$.
This requires $\beta$ to be a vector of length 10 (since $v_j$ goes from
1 to 8, the term $\beta_{v_j}$ goes from $\beta_1$ to $\beta_8$).

Why \eqref{eq:modmat}?  Let us follow the logic of monotonicity and see
the consequences.  We play a trick commonly used
with all forms of regression models (LM, GLM, and aster):
imagine an arbitrary individual (not necessarily in the given data)
having the same true unknown parameter vector $\beta$
but arbitrary covariate value $x$.
We can take this individual to have
graph \eqref{gra:one} in which case the variable indices are 1 to 8 and
expected fitness is
\begin{equation} \label{eq:sigh}
   \psi(x) = \sum_{j = 5}^8 \mu_j(x),
\end{equation}
where we have written the mean value parameters as a function of the covariate
$\mu_j(x)$.  The function $\psi(x)$
is the fitness landscape (for this aster model); it gives the fitness of one
(hypothetical) individual as a function of that individual's phenotypic
covariate value $x$.

Even though we have an explicit formula for $\mu_j(x)$ as a function of $x$,
it is so messy as to be useless for reasoning about the fitness landscape.
So we use the multivariate monotonicity property \eqref{eq:multi-mono-mu}.
For any two distinct values $x$ and $x^*$ of the covariate
\begin{align*}
   0
   & <
   \sum_{j = 1}^8 [ \mu_j(x) - \mu_j(x^*) ] [ \varphi_j(x) - \varphi_j(x^*) ]
   \\
   & =
   \sum_{j = 5}^8 [ \mu_j(x) - \mu_j(x^*) ] [
   ( \beta_9 x + \beta_{10} x^2 )
   -
   ( \beta_9 x^* + \beta_{10} (x^*)^2 ) ]
   \\
   & =
   [ ( \beta_9 x + \beta_{10} x^2 ) -
   ( \beta_9 x^* + \beta_{10} (x^*)^2 ) ]
   \sum_{j = 5}^8 [ \mu_j(x) - \mu_j(x^*) ]
   \\
   \\
   & =
   [ ( \beta_9 x + \beta_{10} x^2 )
   -
   ( \beta_9 x^* + \beta_{10} (x^*)^2 ) ]
   [ \psi(x) - \psi(x^*) ]
\end{align*}
where the first equality comes from plugging in \eqref{eq:modmat} and realizing
that the $\beta_{v_j}$ terms cancel because they do not involve $x$ and
that the $j = 1$, 2, 3, and 4 terms are zero because for them $u_j = 0$.
This implies that both factors in the final expression must have the same
sign, hence
\begin{equation} \label{eq:conclusion}
   \beta_9 x + \beta_{10} x^2 < \beta_9 x^* + \beta_{10} (x^*)^2
   \quad \text{if and only if} \quad
   \psi(x) < \psi(x^*)
\end{equation}
We express \eqref{eq:conclusion} in words by saying
that fitness on the canonical parameter scale $\beta_9 x + \beta_{10} x^2$
is a monotone function of expected fitness $\psi(x)$ and vice versa.
Although we have only given the argument for one concrete case,
the same argument works when $\beta_9 x + \beta_{10} x^2$ is replaced by any
function of any number of covariates \citep[appendix on monotonicity]{aster3}.

This monotonicity relationship is the key to constructing scientifically
interpretable aster models.  We must model on the canonical parameter scale
for many technical reasons: to avoid explicit restrictions on the parameters,
to have concave log likelihood for which optimization is simple, to have
the properties discussed in Sections~\ref{sec:sufficient} and~\ref{sec:entropy}
below.  For scientific interpretability, however, we must be able to interpret
our models on the mean value parameter scale.  If there were no understandable
relationship between the two scales --- just an arbitrary function with no
simple properties --- then there would be no way to interpret our models.

There is a simple consequence of multivariate monotonicity that is sometimes
presented in its place \citep[discussion section]{gws}: if one canonical parameter is increased (the rest being held fixed), then the corresponding mean value
parameter is increased (the other mean value parameters change too but can go
any which way).  This statement does not contain the full content of
multivariate monotonicity and cannot replace it for mathematical derivations,
but it can give some idea of its import without going into too much mathematics.
For canonical linear submodels, this becomes: if $\beta_k$ is increased (other
betas being held fixed), then the expectation of the $k$-th submodel canonical
statistic, the $k$-th component of $M^T y$ increases (expectations of other
components of $M^T y$ change too but can go any which way).  Thus even naive
users should know about the submodel canonical statistic vector $M^T y$.

\subsection{Sufficiency} \label{sec:sufficient}

\citet{fisher} introduced not only the method of maximum likelihood but also
the notion of sufficiency.  A statistic (function of the data that does not
involve parameters) is \emph{sufficient} if the conditional distribution of
the data given the statistic does not depend on the parameter, that is,
for data $y$ a statistic $t(y)$ is sufficient if the conditional distribution
$$
   f_\theta\bigl(y \mid t(y)\bigr)
$$
does not actually depend on the parameter $\theta$.  This means that the
distribution of $y$ given $t(y)$ tells nothing about the parameter.  Hence
any estimate of the parameter should only involve the data $y$ through the
sufficient statistic $t(y)$.  This is the principle of sufficiency: any
sensible estimate of the parameter must be a function of the sufficient
statistic.

The definition of sufficiency is hard to apply, but there is a much simpler
criterion that is much easier.  If the log likelihood is a function
of the data only through a statistic $t(y)$, then that statistic is sufficient.
An exponential family only involves the data through the canonical statistic,
hence the canonical statistic is always sufficient.  Also the method of maximum
likelihood automatically obeys the sufficiency principle, since the maximizer
of the likelihood only involves the data through the log likelihood which
in turn only involves the data through the sufficient statistic.  Thus
sometimes the canonical statistic of an exponential family is called
the canonical \emph{sufficient} statistic, although the ``sufficient'' is
redundant (every canonical statistic is sufficient).

Sufficiency may seem a pedantic distinction, but before \citet{fisher},
no one had identified this notion nor were there any methods of estimation
in common use that obeyed the principle of sufficiency.  Hence every
methodology introduced before 1922 was either (in hindsight) wrong or was
a special case of Fisher's methods (in particular, the method of least squares
is a special case of the method of maximum likelihood if the error distribution
is assumed to be homoscedastic normal).

The principle of sufficiency applied to aster models motivates unconditional
canonical linear models ($\varphi = M \beta$).  In such models the (submodel)
canonical statistic vector $t(y) = M^T y$ is sufficient.  Your humble author
is of the opinion that in elementary statistics teaching too much is made
of the linear change-of-parameter
\begin{subequations}
\begin{equation} \label{eq:change-of-parameter}
   \varphi = M \beta
\end{equation}
and too little is made of the linear change-of-variable
\begin{equation} \label{eq:change-of-variable}
   t(y) = M^T y
\end{equation}
\end{subequations}
from data to (submodel) sufficient statistic.\footnote{In the example in the
preceding section, the concrete expression of \eqref{eq:change-of-parameter}
is \eqref{eq:modmat}.  The concrete expression of \eqref{eq:change-of-variable}
was not derived, and we do so now.  The bilinear form in the log likelihood is
$$
   \inner{y, M \beta}
   =
   \sum_{j \in J} y_j \bigl[ \beta_{v_j} + u_j (\beta_9 x_j + \beta_{10} x_j^2)
   \bigr]
$$
and the coefficient of each beta is the corresponding canonical statistic.
Thus the canonical statistics are
\begin{align*}
   t_k(y) & = \sum_{\substack{j \in J \\ v_j = k}} y_j, \qquad k = 1, \ldots, 8
   \\
   t_9(y) & = \sum_{j \in J} y_j u_j x_j
   \\
   t_{10}(y) & = \sum_{j \in J} y_j u_j x_j^2
\end{align*}
These ten statistics are the components of the sufficient statistic vector.
``Observed = expected'' says the method of maximum likelihood matches the
observed and expected values of these ten statistics.  The sufficiency
principle says all inference should depend on the data only through these
ten quantities. \label{foot:land}
}
In fact, the latter is not
mentioned at all in most courses about LM and GLM.  Clearly each determines
the other, because each determines and is determined by the model matrix $M$.
The question
is: which one does one choose to motivate the choice of $M$?
Most courses about LM and GLM concentrate on \eqref{eq:change-of-parameter},
teaching students how to interpret this equation.\footnote{More precisely,
they teach students how to interpret concrete instances of this equation,
\eqref{eq:modmat} for example.  The abstraction $\varphi = M \beta$ may
never be mentioned in lower level courses in which students are not expected
to understand matrices.}
Since \eqref{eq:change-of-variable} is never mentioned in most courses,
students do not learn to use it to motivate $M$.

This is wrongheaded.  Data are much more concrete than parameters.  Statistical
models are very abstract, and parameters even more so, since parameters are
a mere index for models (parameters can be changed without changing the model,
the family of probability distributions indexed).  We sometimes express this
by the slogan \label{pg:meaningless}
\begin{quote}
Parameters are meaningless quantities.  Only probabilities and expectations
have scientific meaning.
\end{quote}
Of course, some parameters are expectations.  In aster models $\xi$ and $\mu$
have scientific interpretation tied directly to data; they are expectations
(conditional or unconditional) of components of the data (more precisely,
the components of $\xi$ have the relationship to conditional expectations
recalled in footnote~\ref{foot:xi}).  Other parameters
have no such meaning.  In aster models $\beta$, $\theta$, and $\varphi$ are
only weakly tied to data, through the monotonicity properties discussed at
the end of the preceding section.

These considerations say that \eqref{eq:change-of-variable} has a much more
direct connection to data and hence to reality
than \eqref{eq:change-of-parameter}.  It is a triumph of abstraction that
\eqref{eq:change-of-parameter} is preferred to \eqref{eq:change-of-variable}
in elementary statistics teaching.

Aster models are more directly tied to reality
when \eqref{eq:change-of-variable} is used to motivate the choice of model
(by choice of model matrix $M$): the principle is to choose $M$ so that
the components of the (submodel) canonical statistic vector $M^T y$ have
sensible scientific interpretation.  Then the aster model has sensible
scientific interpretation.  The ``observed = expected'' property assures
that the MLE of the (submodel) mean value parameter vector $M^T \hat{\mu}$
also has sensible scientific interpretation,
since it makes $M^T y = M^T \hat{\mu}$.

We call the map $t$ defined by \eqref{eq:change-of-variable} the
\emph{sufficient dimension reduction} map.  It reduces the dimension of the
data vector $y$ to a linear function $M^T y$ of smaller dimension
(the dimension of the parameter $\beta$) \emph{without losing information}
because $M^T y$ is sufficient.

Sufficiency is a very abstract property.  It is a property of a statistic
relative to a statistical model.  Throwing away the all of the data except
for the sufficient statistic vector loses no information \emph{about that
particular statistical model}.  The point of sufficiency is purely
philosophical.  And it works both ways.  Given a model, one wants to know
the sufficient statistics.  Given a set of scientifically important statistics,
one wants to choose a model for which those statistics are sufficient.  So
in aster models, we want to choose canonical linear submodels in which the
sufficient statistic vector $M^T y$ has components that are important
scientifically interpretable quantities.

For a specific instance of this reasoning, let us return to the estimation
of fitness landscapes introduced in the preceding section.
In footnote~\ref{foot:land} we derived the submodel canonical sufficient
vector.  An earlier methodology for estimating fitness landscapes, or, more
precisely, the best quadratic approximation (BQA) to fitness landscapes, was
introduced by \citet{la}.  This method uses LM rather than aster to estimate
the fitness landscape.  If fitness satisfied the assumptions for LM, that
the conditional distribution for fitness given phenotypic trait covariate
values is homoscedastic normal, then the sufficiency argument applied to that
model would give the same sufficient dimension reduction to the same ten
sufficient statistics $t_k(y)$, $k = 1$, $\ldots$, 10 that are the sufficient
statistics for the aster model.  This is because the two models agree about
what fitness is (as a function of the data $y$), because they both use
a quadratic model, and because both are exponential family models.
So aster modelers agree completely with \citet{la} about the sufficient
dimension reduction map.  The only disagreement is about the statistical model:
the LM used by \citet{la} is often grossly wrong for fitness landscapes
whereas aster models may have no apparent wrongness.

\subsection{Efficiency} \label{sec:efficiency}

\citet{fisher} introduced not only the method of maximum likelihood and the
notion of sufficiency but also the notion of efficiency.  The efficiency
of an estimator $\hat{\varphi}$ of a scalar parameter $\varphi$ is the expected
squared error
\begin{equation} \label{eq:squared-error-scalar}
   E \{ (\hat{\varphi} - \varphi)^2 \}.
\end{equation}
The smaller the expected squared error, the more efficient the estimator.
Here $\hat{\varphi}$ is a function of the data, hence a random variable,
hence it has a distribution,\footnote{In all but the simplest cases we do
not have an explicit formula for the distribution, but we know it exists
because every random vector has a distribution.  This distribution, called
the \emph{sampling distribution} of the estimator $\hat{\varphi}$ is a
sticking point many people have with statistics.  You have only one
data set, and you have only one estimator $\hat{\varphi}$ calculated from it.
It takes an act of imagination to visualize the distribution of possible other
values of the estimator that would result from imaginary replications of the
experiment or observational study and to think about properties of
this distribution, such as efficiency.}
and it is with respect to this distribution
that the expectation \eqref{eq:squared-error-scalar} is defined.
In contrast, the parameter $\varphi$
is considered nonrandom, a property of the probabilistic law of nature that
governs the generation of the data.  The estimator $\hat{\varphi}$, being
random, would be different if the experiment or observational study were
repeated, so the amount of error $\hat{\varphi} - \varphi$ is a random
quantity, and we cannot say how large it would be, only how large it
is on average.
Hence the definition of efficiency.  A more efficient estimator provides
more accurate estimates, on average.

When the parameter is a vector $\varphi$,
then so is the estimator $\hat{\varphi}$,
and the definition of efficiency becomes more complicated,
it is
\begin{equation} \label{eq:squared-error-vector}
   E \{ (\hat{\varphi} - \varphi) (\hat{\varphi} - \varphi)^T \},
\end{equation}
which denotes the matrix whose $j, k$ component is
$E \{ (\hat{\varphi}_j - \varphi_j) (\hat{\varphi}_k - \varphi_k) \}$.
Having a matrix measure of efficiency is more difficult to understand.
If two estimators have efficiency matrices $V_1$ and $V_2$, then we say
the first is more efficient than the second if $V_2 - V_1$ is a positive
semi-definite matrix.  It may be that neither $V_2 - V_1$ nor $V_1 - V_2$ is
positive semi-definite, and this means that some but not all components
of one estimator
are more efficient estimates of the corresponding parameter than those of
the other estimator (neither estimator is better for all components).

The exact efficiency of an estimator is usually impossible to calculate
when the estimator is a complicated function of the data.  Thus we make
do with the so-called ``asymptotic'' distribution, which refers to the limit
as the number of individuals in the data set goes to infinity.
Under certain conditions the asymptotic distribution of the MLE $\hat{\varphi}$
is known to be normally distributed,
and this (multivariate) normal distribution is centered
at the true unknown parameter vector $\varphi$, and has variance matrix that
is the inverse of the matrix
$$
   I(\varphi) = E_\varphi\{ - \nabla^2 l(\varphi) \},
$$
which is called the \emph{Fisher information matrix}, because it too was
introduced by \citet{fisher}.
For a general exponential family of distributions, and in particular for
a saturated aster model, this simplifies to
$$
   I(\varphi) = \nabla^2 c(\varphi).
$$
For a canonical linear submodel with parameterization $\varphi = M \beta$,
this simplifies to
\begin{equation} \label{eq:fisher-information}
   I(\beta) = M^T \nabla^2 c(M \beta) M
\end{equation}
\citep[Section~3.2]{gws}.
So for exponential families of distributions in general and aster models
in particular we know that the asymptotic distribution of the MLE
$\hat{\beta}$ is normal with mean $\beta$ (the true unknown parameter value)
and variance $I(\beta)^{-1}$ (inverse Fisher information).
The \emph{asymptotic efficiency} is this asymptotic variance of the MLE
$I(\beta)^{-1}$.  The asymptotic normal distribution of other parameters
can be derived via the delta method \citet[Section~3.3]{gws}.

It is a remarkable fact about statistical estimation that maximum likelihood
estimates are as efficient as it is possible for an estimator to be.  This
fact was also first surmised by \citet{fisher}.  It is also referred to as
the \citeauthor{cramer}-Rao lower bound.\footnote{Rao was a student of Fisher.
\citet{cramer} was the first recognizably modern textbook of theoretical
statistics and included perhaps the first truly rigorous proof of the
asymptotic normality of the MLE.}

This optimality
property of maximum likelihood estimators (they are most efficient) is not
quite correct without conditions.  A large part of the mathematical statistics
in the twentieth century was devoted to proving asymptotic normality of MLE
under weaker and weaker conditions and to weakening the conditions under
which MLE were efficient. \Citet[Sections~5.3, 5.5, 8.5, and~8.6]{vdv} gives
the current state of the theory (but is not easy to read).

Asymptotic efficiency is a very abstract property.  It is a property of an
estimator (such as the MLE) relative to all other estimators.  It says that
an asymptotically efficient estimator (such as the MLE) is at least as good
as any other estimator for sufficiently large sample sizes (as $n$ goes to
infinity) when squared error \eqref{eq:squared-error-scalar} or
\eqref{eq:squared-error-vector} is the criterion.  And why that criterion?
Partly mathematical convenience --- it is easy to work with and leads to
interesting results --- and partly the connection with asymptotic normality
--- for the multivariate normal distribution, the variance matrix is the
natural criterion of error because it is a parameter of the distribution.
The point of asymptotic efficiency is purely philosophical.  It justifies
using estimators (such as the MLE) that are efficient.

\subsection{Entropy} \label{sec:entropy}

Boltzmann made the connection between entropy and probability.
He considered this his most important discovery, having $S = k \log W$
engraved on his tombstone ($S$ is entropy, $W$ is probability, and $k$
is a physical constant, now known as Boltzmann's constant).

Exponential families have a simple relationship with entropy \citep{jaynes}.
The
entropy of one probability distribution with probability function $f(y)$ with
respect to another with probability function $g(y)$ is
\begin{equation} \label{eq:entropy}
   E_f\left\{ \log \frac{f(y)}{g(y)} \right\}
   =
   \sum_y f(y) \log \frac{f(y)}{g(y)},
\end{equation}
where the sum runs over allowed values of $y$ (we consider here only the
case where $y$ is discrete, but the argument when some components of $y$
are continuous is similar).
Consider the following somewhat unmotivated problem (it is interesting that
this setup leads to an interesting answer but the setup is not interesting
in itself).  We wish to maximize \eqref{eq:entropy} subject to the constraints
\begin{equation} \label{eq:constraints}
   \tau_k = E_f\{ t_k(y) \} = \sum_y f(y) t_k(y), \qquad k \in K
\end{equation}
where the $\tau_k$ are arbitrary real numbers and the $t_k$ are arbitrary
real-valued functions, and where $f$ is allowed to vary
over all possible real valued functions that specify
probability distributions, hence satisfy the additional constraints
$f(y) \ge 0$ for all $y$ and $\sum_y f(y) = 1$.  The solution to this problem
has the following form
$$
   f(y) = g(y) \exp\left( \sum_{k \in K} \beta_k t_k(y) - c(\beta) \right),
$$
where the $\beta_k$ are Lagrange multipliers (real numbers chosen
to make the constraints \eqref{eq:constraints} hold, $\beta$ is the vector
having components $\beta_k$, and $c(\beta)$ is a function determined by
the requirement that the probabilities sum to one.
Introduce the vector $t(y)$ having components
$t_k(y)$ and this solution can be rewritten
$$
   f(y) = g(y) e^{\inner{t(y), \beta} - c(\beta)}
$$
a family of distributions (if we consider $\beta$ a parameter vector)
which is an exponential family because the log likelihood is
$$
   \log f(y) = \inner{t(y), \beta} - c(\beta)
$$
(we can drop the term $\log g(y)$ from the log likelihood because it does not
contain the parameter).  The canonical statistic vector is $t(y)$ and the
canonical parameter vector is $\beta$.

This discussion is a little odd.  We started with the idea that the
constraint equations \eqref{eq:constraints} fixed the values of certain
expectations, but now we are considering $\beta$ a parameter, which gives
different distributions $f(y)$ as $\beta$ varies over its allowed values.
We know from exponential family theory that the relationship between
the canonical parameter vector $\beta$ and the mean value parameter vector
$$
   \tau = E_\beta\{ t(y) \}
$$
is invertible (each vector $\beta$ corresponds
to a distinct vector $\tau$).  Thus there is at most one distribution in the
family that solves the constraints \eqref{eq:constraints}.  But having
gotten to this point we now drop the idea that we were attempting to fix
the mean vector $\tau$ at just one value.  If we allow $\tau$ to vary over
its allowed values we get the whole exponential family because (to repeat
what was just said) the map $\beta \to \tau$ is invertible so either $\beta$
or $\tau$ can serve as a parameter indexing the family.

Thus we have arrived at a conclusion that can be somewhat sloppily stated
as saying that exponential families arise by maximizing entropy subject to
fixing the means of a set of quantities, which turn out to be the canonical
statistics of the exponential family that arises from entropy maximization.
If one chooses a set of statistics, then the family of probability
distributions that maximizes entropy is the exponential family having
these chosen statistics as its canonical statistics.

We now emphasize that the distribution $g$ which has been fixed throughout
the argument plays a role too.  Each different choice of $g$ gives a different
exponential family.  In the context of aster models we let $g$ be
the probability function of any distribution in the saturated
model, which is given by biological considerations that lead to a particular
graphical model.  The maximum entropy argument applies to submodels.
If we choose statistics $t(y)$ of the form $t(y) = M^T y$, in which case
the vector of means being fixed is $\tau = M^T \mu$,
then we know this does not take us outside the original saturated model
but rather leads to a canonical linear submodel.  The maximum entropy
argument then justifies these submodels saying that they are the models
that maximize entropy subject to fixing the values of the means
of the submodel canonical statistic vector $M^T y$.

Even more sloppily, we can say that the maximum entropy argument says that
if we choose the submodel canonical statistic vector $M^T y$ to be
scientifically interpretable, then the corresponding canonical linear submodel
will also be scientifically interpretable.  It is the model that maximizes
entropy (randomness) subject to controlling the mean values of $M^T y$.

Maximum entropy is a very abstract property.  It is a property of a family of
statistical models (each a family of probability distributions) relative to
a statistic vector.  It says that the one family in the family of families
that is an exponential family having the chosen statistic vector as its
canonical statistic vector maximizes entropy (within the family of families).

So why maximize entropy?
All physical processes maximize entropy.
Thus we expect maximum entropy models with canonical statistics that are
scientifically important quantities to model the biological situation well.
The point of maximum entropy is purely philosophical.  It justifies
using exponential family statistical models and focuses attention on their
submodels that are themselves exponential families (the canonical linear
submodels).

\section{Discussion} \label{sec:discuss}

\begin{quote}
All models are wrong, but some are useful. \\ \hspace*{\fill} --- George Box
\end{quote}

The important philosophical point about all of this is how it all hangs
together.  Many different bits of mathematical statistics play a role.
All are part of aster models being a practical solution to biological
problems.

Start with the fact that aster models are statistical models.  One does not
need a statistical model to do some simple forms of statistical inference,
but when one
has a model --- a joint probability distribution for all the data expressed
as a function of parameters of the model --- the full panoply of statistical
methodology is available.  If one has a statistical model, then \emph{any
question whatsoever} that can be phrased in terms of probabilities and
expectations can be, in principle, answered by calculations based on the model.
In practice, exact calculation may not be possible but approximate calculation
based on simulation of the model is always possible.

In life history analysis, as recounted in the introduction of \citet{aster2},
before aster models there was no way to construct a statistical model for
complete life history data (at least no ideas were proposed before aster
models).  In the absence of statistical models for complete data, many
approaches were devised.  One approach was to separately analyze parts
of the data called ``components of fitness'' to which available models
could be applied.  But there was no way to combine these analyses to
make inference about fitness itself.  Other approaches analyzed fitness
itself without a valid model.  Both demographic approaches using Leslie
matrices and Lande-Arnold analysis \citep[see][for references]{aster2}
tacitly assume a normal
distribution for fitness, which is often grossly and obviously wrong.
All these approaches
were recognized to be unsatisfactory and criticized in the literature
reviewed by \citet{aster2}, but they were used despite their
known problems because nothing else was available before aster came along.

The title of \citet{aster2} is ``\emph{Unifying} life history analysis
$\ldots$'' (our emphasis).  Previously existing approaches to
life history analysis could be applied, even unsatisfactorily, to only
a small fraction of the analyses to which aster models can be applied.
None could do more than one of the three examples
of aster analysis in \citet{aster2}, and none could do as well as aster.

So the first virtue of aster models is to provide a unified theory of
defensible statistical models for complete life history data.
The description ``defensible'' is
used advisedly.  All statistical models are wrong.  None can incorporate
all the biology relevant to the data.  But a model that is not obviously
grossly wrong --- one that passes statistical tests of goodness of fit ---
can be useful, as the Box aphorism says.

Having a defensible statistical model may be enough for sensible statistical
inference in simple situations, but aster models are too complicated for that.
Having a joint statistical model for the complete data does not, by itself,
avoid the ``components of fitness'' problem.  One must still use a model
complicated enough so that all components of fitness are modeled well.
In an aster model, such a model will typically have too many parameters to
estimate well if the original (conditional canonical) parameterization is used.
And even if there are not too many parameters, those parameters have no
simple relationship with unconditional means, so it is difficult to choose
submodels and difficult to interpret any deemed to fit.

It is the change-of-parameter given by equations \eqref{eq:theta-to-phi}
from conditional to unconditional canonical parameters that is the second
virtue of aster models.  Because of the monotonicity, sufficiency, and maximum
entropy properties of exponential families, these parameters, when modeled
linearly \eqref{eq:canonical-linear-submodel} lead to models that have
canonical sufficient statistics that are scientifically interpretable and
and have mean value parameters connected by multivariate monotone relationship
(Section~\ref{sec:moron}) with the canonical parameters that are modeled.
It is this change-of-parameter that allows complicated data to be adequately
modeled with few enough parameters to be estimated well, and it is this
change-of-parameter that allows scientific interpretation of these models.

The change-of-parameter from conditional to unconditional canonical parameters
has a very abstract origin.  For each concrete instance of an aster model ---
a saturated model determined by a graph with annotation and a submodel of it
determined by a model matrix --- this change-of-parameter could be given
by explicit formulas like \eqref{eq:big-theta-to-phi}, which whether explicitly
written out or not are ``understood'' and used by the aster software,
but no one would invent such a change-of-parameter except from theoretical
considerations.  The change of parameter \eqref{eq:theta-to-phi} arises from
the recognition that an aster model is an exponential family and from
the process of finding its canonical parameter vector.
Without the notion of an abstract exponential family,
this change-of-parameter and all the properties of aster models that flow from
it would never occur to anyone.

The change-of-parameter from conditional to unconditional canonical parameters
causes some conceptual difficulties to those trained to think about components
of fitness.  The unconditional aster model directly addresses overall fitness
but does not directly address the separate components (the conditional aster
model does that).  When one wants to have one's cake and eat it too (study
both overall fitness and the separate components), unconditional aster models
can do the job, but only in an indirect way that is difficult to wrap one's
mind around.  Example~2 of \citet{aster2} presents data on fitness of
\emph{Echinacea angustifolia}.  The experiment was done in two parts.
Plants started life indoors in a growth chamber.  Those that survived this
initial period were replanted outdoors in an experimental field plot.
Of interest were differences in fitness for different ``cross types''
representing different amounts of inbreeding.  The unconditional aster model
found there were statistically significant differences among cross types.

So far so good.  The unconditional aster model made the analysis of overall
fitness simple.  Confusion creeps in when the question is asked whether the
fitness differences are due to differential survival in the growth chamber
or differential survival and growth (reproduction was not measured in these
data) in the field plot.  If one adds additional terms to the aster model
modeling fitness differences among cross types both in the growth chamber
and in the field plot, the growth chamber terms are not statistically
significant ($P = 0.34$).  So that says differential survival in the growth
chamber doesn't affect fitness?  No!  In an unconditional aster model terms for
later components of fitness propagate back through the equations to earlier
ones.  This is seen in \eqref{eq:big-phi-to-theta} where the expression for
$\theta_1$ contains all components of $\varphi$,
the expression for $\theta_2$ contains all components of $\varphi$
except $\varphi_1$ and $\varphi_5$, and so forth.  Thus the comparison
between submodels with and without explicit growth chamber terms says that
differential survival in the growth chamber has no effect on fitness over
and above what the maximum entropy principle automatically builds into the
submodel that has only terms for overall fitness.  Confidence intervals
for survival in the growth chamber \citep[Figure~2A]{aster2} show clear
differences among cross types, whether or not the model has
terms for fitness differences among cross types in the growth chamber.

In short, trying to have one's cake and eat it too --- trying to analyze
both overall fitness and separate components of fitness --- causes confusion.
Biologists know that ``components of fitness'' only act together not
separately, so analyzing the effects of the components of fitness separately
must be problematic.  Aster models treat this problematicity correctly: only
overall fitness is modeled simply.  Once one has found a defensible aster
model that passes tests for goodness of fit \citep[in Example~2 of][the model
with only terms for overall fitness differences among cross types]{aster2},
one can look at parameter estimates for various conditional or unconditional
mean values to get some idea of the contributions of the separate components
of fitness but one cannot completely disentangle them.

Aster models are very complicated compared to LM and GLM and very simple
compared to other graphical models.  Why are aster models defined they way
they are?  Why not more simple or more complicated?  Aster models are
complicated enough to encompass most but not all life history data.  The
``predecessor is sample size'' property is the key to aster model theory.
Without it, the log likelihood \eqref{eq:logl-cond} would not be linear in
$y$ (both the $y_j$ and $y_{p(j)}$ appearing linearly),
the joint distribution of
$y$ would not be an exponential family, there would be no unconditional
canonical parameterization, and there would be no multivariate monotone
relationship of parameters with unconditional mean values.  Nor would there
be strictly concave log likelihood, which makes MLE unique and easy to find
by computer optimization algorithms.  So the ``predecessor is sample size''
property does a lot.

But ``predecessor is sample size'' is very restrictive.  Many forms of
dependence that can be imagined for life history data do not fit into this
paradigm.  So why stop where aster models stop?  The strong theoretical
properties of aster models mentioned in the preceding paragraph do many
things but have two main effects: they allow
\emph{simple scientific interpretation}
and \emph{efficient computer implementation}.
The simplicity of interpretation comes from
the multivariate monotonicity relationship of canonical and mean
value parameter values and the sufficiency and maximum entropy
properties.  The efficiency of implementation comes from
simple (for the computer) expressions for
the changes-of-parameter $\theta \leftrightarrow \varphi$,
$\xi \leftrightarrow \mu$, $\theta \to \xi$, and $\varphi \to \mu$,
through concavity properties of log likelihood that guarantee uniqueness
of MLE and allow MLE and changes of parameter $\xi \to \theta$
and $\mu \to \varphi$ to be computed by straightforward computer optimization,
and through the relation \eqref{eq:fisher-information} that gives Fisher
information as a simple (for the computer) expression
and through it the asymptotic distribution of the MLE and resulting
hypothesis tests and confidence intervals.
Any generalization of aster models other than those already
considered in \citet{gws} and mentioned in various footnotes
(\ref{foot:affine}, \ref{foot:group}, \ref{foot:infinitely-divisible},
\ref{foot:barndorff-complete}, \ref{foot:restrict}) in this technical report
would lose (as far as we can see) at least one of the the properties that
give aster models their simple scientific interpretation and efficient
computer implementation.
\begin{itemize}
\item Having defensible statistical models allows likelihood inference.
\item The efficiency property of maximum likelihood estimates guarantees
    they are good.  Without the efficiency of MLE, one would have no idea
    how to proceed with complicated life history data.
\item The relationship between conditional and unconditional mean value
    parameter vectors (Section~\ref{sec:sample-size} above) is peculiar
    to ``predecessor is sample size'' models.
\item The multivariate monotone relationship between canonical and mean value
    parameter vectors (Sections~\ref{sec:intro} and~\ref{sec:moron} above)
    is peculiar to exponential family models.
\item Cumulant functions (Sections~\ref{sec:expo-general}, \ref{sec:expo},
    \ref{sec:geyer-great-idea}, \ref{sec:convex}, \ref{sec:submodel},
    and~\ref{sec:efficiency}) are peculiar to exponential family models.
    They allow means, variances, and covariances of components of the
    canonical statistic vector, hence also the Fisher information matrix
    and the asymptotic distribution of the MLE to be calculated from
    derivatives of cumulant functions, which are simple for the computer
    to calculate.
\item The relationship between IID and exponential families
    (Section~\ref{sec:expo-general}) is peculiar to exponential family models.
    This makes the ``predecessor is sample size'' property lead to exponential
    family joint distributions (Section~\ref{sec:expo}).
\item The sufficient dimension reduction argument (Section~\ref{sec:sufficient}
    is unique to exponential family models.
\item The maximum entropy argument (Section~\ref{sec:entropy})
    is unique to exponential family models.
\item That ``structural zeros'' (footnote~\ref{foot:zeros}) are handled
    automatically and correctly is unique to maximum likelihood estimation
    based on a statistical model.
\end{itemize}
All of these points are involved in making aster models work the way they do.

\section{Statistical Models and Their Interpretation}

We close with a ``post-discussion discussion'' of an important point that often
arises in the context of aster models but really has nothing to do with them,
since it applies to all statistical models.  This issue is reification
of parameters.

Probability models are already very abstract.  The nature of probability
and expectation has been argued about by scientists and philosophers for
hundreds of years and no agreement has been reached \citep{hajek}.
Mathematicians cut
through the confusion by axiomatizing probability and expectation, giving
them purely mathematical definitions cut loose from debates about what features
of the real world might or might not correspond to them.
\citet{kolmogorov-foundations} gave a system of axioms for probability theory
and since then all research-level probability theory been based on them.
Their level of abstraction is very high, requiring the notion of abstract
integration over so-called measurable spaces (abstract sets equipped with
sigma-algebras).  It is so high that no one attempts to teach this theory
to undergraduates or master's students.  This abstract integration theory
is only taught in Ph.~D. level courses in probability, theoretical statistics,
and real and functional analysis (from textbooks like \citealp{fristedt}
and \citealp{rudin}).  Consequently, pre-1933 notions of
probability and expectation are still taught in all universities in courses
designed for undergraduates and master's students
(from textbooks like \citealp{casella}).  Most scientists who use
probability and statistics are exposed to them in such courses.
Call this theory ``master's level probability theory'' to give it a name
(all of the probability theory in this technical report is ``master's level'').

Probability models in ``master's level probability theory'' are still very
abstract (if not quite as abstract as in research level probability theory).
Probability and expectation are still given purely mathematical definitions
cut loose from debates about what features of the real world might or might
not correspond to them.  Discrete probability models are defined by giving
a probability mass function (PMF), which is a real-valued function whose
domain is an arbitrary (abstract) set $S$ and satisfies two properties
\begin{subequations}
\begin{equation} \label{eq:nonnegativity}
   f(y) \ge 0, \qquad y \in S
\end{equation}
(nonnegativity) and
\begin{equation} \label{eq:sums-to-one}
   \sum_{y \in S} f(y) = 1
\end{equation}
\end{subequations}
(sums to one).  Facetiously, one can say that (master's level) probability
theory is the study of functions that are nonnegative and sum to one.
The domain $S$ of the PMF is called the \emph{sample space}.
A real-valued function on the sample space is called a \emph{random variable},
and the \emph{expectation} of a random variable $g(y)$
is given by
\begin{equation} \label{eq:discrete-expectations}
   E\{g(y)\} = \sum_{y \in S} g(y) f(y)
\end{equation}
if the sum exists (which it may not if the sample space is infinite).
Probability is a special case of expectation; it is expectation of Bernoulli
(zero-or-one-valued) random variables.  If $g$ is Bernoulli, define
$$
   A = \set{ y \in S : g(y) = 1 }.
$$
In this case the expectation \eqref{eq:discrete-expectations} becomes
\begin{equation} \label{eq:discrete-probabilities}
   P(A) = \sum_{y \in A} f(y)
\end{equation}
because $g(y) = 1$ for $y \in A$ and $g(y) = 0$ for $y \notin A$.
The sum in \eqref{eq:discrete-probabilities} always exists because it is
less than the sum in \eqref{eq:sums-to-one} which exists by definition of
PMF.  Any subset of the sample space is called an \emph{event}
and \eqref{eq:discrete-probabilities} is called
the \emph{probability of the event $A$}.

``Master's level probability theory'' also studies continuous probability
models in which the sums above are replaced by integrals, but we need not
discuss them here, because all aster models that have been used in published
papers are discrete, so that is all the probability theory we need for this
discussion.

Notice that, as we said above, these definitions have not the faintest trace
of the philosophical arguments about what probability and expectation ``really
are.''  Mathematicians have simply declared that anyone's notions
of probability and expectation that do not agree with these formal
definitions \eqref{eq:discrete-expectations}
and \eqref{eq:discrete-probabilities} will not be entertained,
and conversely any notions that do agree with these formal definitions
are fine with them.  Mathematicians do not care about the philosophical
arguments; they define expectation and probability mathematically and get
on with doing mathematics.

A \emph{statistical model} is a family of probability models.  The problem
of \emph{statistical inference} is, given data that is supposed to be from
one probability model in a statistical model, to say something about which
one it is.\footnote{This ``saying something about'' can take the form of
point estimates, such as maximum likelihood estimates, confidence intervals,
or tests of statistical hypotheses.\label{foot:infer}}
This brings us to deep philosophical waters.  Statistical
inference is the statistician's answer to the philosopher's questions of
induction and epistemology.  When in a totalizing mood, statisticians can
think that all learning about the world is statistical
inference.\footnote{This is obvious nonsense because people learned things
long before anything was known of statistics, but there is a grain of truth
in it.  Except for logical tautologies, all knowledge is uncertain, and all
uncertainty can be modeled by probability theory --- whether this gives a
satisfactory and complete description of uncertainty is debatable, but such
modeling is possible --- from which it follows that all learning is statistical
if the uncertainty of knowledge is to be properly accounted for.  This is understood in the branch of philosophy called ``Bayesian epistemology''
\citep{talbott}; our point here is that non-Bayesian statistical inference
also answers epistemological questions.   Most people
most of the time feel no uncertainty about most of their knowledge; the
customs of their tribe, sect, and clique are not questioned.  This goes for
intellectuals and academics too.  Even scientists use statistics only when
the uncertainty of inference is so glaring that any attempt at avoiding
statistics must be forlorn.  They use it only when they have to.  The fact
that in real life not all learning is statistical is a deep mystery.  Most
people most of the time are quite certain about many very questionable things,
and this seems to work.  But why?}  But leaving that aside, what theoretical
statistics is all about is statistical inference, so here we take the idea for
granted.

Statistical models are often specified
by \emph{parameterization}.\footnote{Statistical models that are too big
to be specified by a finite set of parameters exist and are called
``nonparametric.''}  Each probability model in the statistical model is
specified by a PMF that is a function of another variable (or variables)
called the \emph{parameter} (or \emph{parameters}).
When there are multiple parameters, these
are thought of as a single mathematical object called the \emph{parameter
vector}.  The parameterization is indicated by subscripts $f_\theta$ on
the PMF, where $\theta$ is the parameter or parameter vector
(it is a convention to use roman letters for ordinary variables and random
variables and greek letters for parameters).  A subscript indicating the
parameter is also added to the notation for expectation
\begin{equation} \label{eq:discrete-expectations-parmeter}
   E_\theta\{g(y)\} = \sum_{y \in S} g(y) f_\theta(y)
\end{equation}
and probability
\begin{equation} \label{eq:discrete-probabilities-parameter}
   P_\theta(A) = \sum_{y \in A} f_\theta(y).
\end{equation}

Statistical models are also very abstract.  Firstly, they are abstract
because probability models are abstract and each statistical model is
an abstract set whose elements are probability models.  Secondly, there
is not only the abstraction of probability models (ignoring what probability
and expectation ``really mean'') but also the abstraction of statistical
inference.  Statisticians talk about statistical models by saying one
particular parameter value among the abstract set of allowed parameter values
is the \emph{truth} or \emph{true unknown parameter value}, the latter to
emphasize that which is the truth is unknown and the problem of statistical
inference is to say something about which parameter value
it is (footnote~\ref{foot:infer}).  It is ``assumed'' that
if $\theta$ is the truth, then the PDF $f_\theta$ corresponding to it
is the probability model ``for'' the observed data, meaning that, whatever
philosophical attitude we have about the correspondence of probability models
to the real world, we have concluded that this model describes the natural
process generating these data.\footnote{This ``assumed'' is for the sake of
argument, to get on with the mathematics of statistical inference.  The
assumption may later be questioned based on the inferences made.  In their
example~3, \citet{aster2} first ``assume'' an aster model for the data but
later check its validity using residual analysis, their Figure~4.}
Piled on top of the probabilist's abstraction
is the statistician's abstraction: learning the truth (about nature) is
saying something about a parameter in a statistical model.  All learning
can be formalized this way, and statisticians do so.

It is an observable fact about statistics teaching that the concept
of statistical models is considered problematic and completely avoided
in introductory courses.  The terms \emph{statistic} and \emph{parameter}
are much discussed (see footnote~\ref{foot:intro}),
but the abstract notion of a family of probability
distributions indexed by the parameter (so each particular value of the
parameter vector specifies one probability model) is absent.  The distinction
is clearly drawn between a parameter $\theta$ and a statistic $\hat{\theta}$
used to estimate it.  But what is left unmentioned is the fact that $\theta$
specifies the probability model through its PMF $f_\theta$ ---
most introductory statistics textbooks do not develop enough probability
theory to make this clear --- and hence the estimator $\hat{\theta}$ also
specifies a probability model through its PMF $f_{\hat{\theta}}$.

The two probability models are different: statistics are not parameters
so $\hat{\theta}$
is not $\theta$, hence $f_{\hat{\theta}}$ is not $f_\theta$.  Nevertheless,
$\theta$ being unknown, our only guide to what $f_\theta$ says about the data
is $f_{\hat{\theta}}$.  So we use $f_{\hat{\theta}}$ in place of $f_\theta$,
not blindly or stupidly but cautiously and sophisticatedly, worrying about
and quantifying the differences between our calculations based
on $f_{\hat{\theta}}$ and what those same calculations would be if based on
$f_\theta$ (which they ought to be but cannot be because $\theta$ is unknown).

This caution gives rise to an infinite regress: when we worry
about the differences between $f_\theta$ and $f_{\hat{\theta}}$ we realize
those differences are random because $\hat{\theta}$ is random (being a function
of the data, which are random).  Thus we must do probabilistic calculations
based on a PMF, which ought to be $f_\theta$, but we must
use $f_{\hat{\theta}}$ because $\theta$ is unknown.
So we have higher-order worries that our calculations about the
errors made in using $f_{\hat{\theta}}$ instead of $f_\theta$ are themselves
wrong because of the same issue.  So we repeat the process, attempting to
quantify our higher-order errors, and are faced with yet higher-order errors,
and so forth ad infinitum.  Fortunately, the errors get smaller as both order
and sample size
increase and after some point can be ignored.  We raise this infinite regress
to show that statistical inference has some very deep philosophical issues.

But that infinite regress is not the main subject of this section, the main
subject is that most users of statistics, including most scientists, have only
the foggiest ideas about statistical models.  Perhaps this is the fault of
bad statistics teaching or of human discomfort about
randomness, probability, and statistics.  Most scientists do not think of
parameters $\theta$ as mere indices that specify a probability model through
its PMF $f_\theta$ and do nothing more.  Rather they reify parameters as
``facts of nature'' asking what the parameters ``really mean'' not
understanding our dictum (page~\pageref{pg:meaningless}) that parameters are
meaningless in themselves and only have meaning through $f_\theta$ and
probabilities and expectations calculated using $f_\theta$.  If I had a nickel
for every time I've tried to explain that fitting statistical models to data
does not directly discover truth about nature,\footnote{Am I not being
contradictory?  Does statistical inference discover truth or not?  When I
am wearing my statistician hat, yes it does, ``truth'' being \emph{defined}
as the parameter value that indexes the correct probability model,
which is ``assumed'' to be a member of the statistical model being used.
When I am wearing my scientist hat, no it does not.  All models are wrong;
none capture all aspects the phenomena being studied.  Moreover, to the
extent that they capture any aspects, they do so through probabilities and
expectations calculated using the model.  Hence our emphasis on mean value
parameters corresponding to scientifically interesting random variables.  Those
have direct scientific interpretation.  Parameters, in general, do not.}
and that the parameters
estimated can only be interpreted through the statistical model and its
properties, I'd be rich.  I can make the same explanation over and over to
the same person and get the response over and over ``yes, I know that''
but ``these sorts of explanations'' (reifying parameters)
``are what's expected.''  That reifying parameters is just wrong
is never directly confronted.

Returning to aster models, \citet{gws}, \citet{aster2}, and \citet{aster3}
are careful to avoid reifying parameters, so much so that
a referee of \citet{aster3} complained that more explanation of ``what the
parameters mean'' was needed.  When one understands that the parameters
don't ``mean'' anything and that interpretations should refer to the
probability models indexed by parameters rather than the parameters themselves,
how does one provide valid scientific interpretations of statistical results?
One way is hypothesis tests of model comparison (done by the function
\texttt{anova} for aster model fits).  Models have scientific interpretations,
hypothesis tests compare two models, one a submodel of the other, and show
either that the submodel fits as well as the supermodel (so the supermodel
adds nothing useful and scientific interpretation should be based on the
submodel) or it does not (so only the supermodel fits the data and scientific
interpretation should be based on it).  Often, this is all that needs to be
done: report which model fits the data best.  When one wants more, an
explanation of what this ``best'' model does and does not say, one must
be very careful, as we saw in the discussion of Example~2 of \citet{aster2}
(Section~\ref{sec:discuss} above).  What \citet{aster2} did in that example
was use confidence intervals for mean value parameters as the basis of
their scientific interpretation.  This was also done in the example in
\citet{gws}.  In Example~1 of \citet{aster2} the parameter of scientific
interest was the population growth rate $\lambda$ which is a function of
the mean value parameter vector $\mu$ through the ``stable age equation''
so the principle of invariance of maximum likelihood estimates
(footnote~\ref{foot:invariance}) can be used to find the MLE of $\lambda$
and the delta method can be used to find a confidence interval for $\lambda$.
In Example~3 of \citet{aster2} and the examples of \citet{aster3} the object
of scientific interest is the fitness landscape, like the function $\psi$
defined by \eqref{eq:sigh}.  This is an infinite-dimensional parameter
(a parameter vector with an infinite number of components),
being a whole function and the components
being the values of the function for each possible argument value.
So it can
only be ``reported'' by plotting a graph of the function, which these papers
do.  Confidence regions for infinite-dimensional parameter vectors are
possible, but too complicated to be of much use; none are reported in these
papers.  Instead, confidence regions are reported for specific aspects of
this function such as the point where it achieves its maximum (the fitness
optimum).  Seen in this light, scientific interpretation of
statistical results is tricky.  The scientific interpretation of a statistical
model is by no means obvious.  One must decide which probabilities of what
events,
expectations of what statistics, hypothesis tests comparing what models,
or confidence intervals for what parameters should be computed to provide
a clear and convincing scientific interpretation.  The job is only started
when the ``best'' statistical model is found.

\section*{Acknowledgments}

The impetus for this technical report was helping a philosopher,
Emily Grosholz, get up to speed on aster models in aid of writing a paper
on them \citep{grosholz}.  But the process went both ways.  Questions and
comments from Emily led to large improvements in this technical report.

\begin{thebibliography}{}

\bibitem[Barndorff-Nielsen(1978)]{barndorff}
Barndorff-Nielsen, O.~E. (1978).
\newblock \emph{Information and Exponential Families}.
\newblock John Wiley, Chichester.

\bibitem[Brown(1986)]{brown}
Brown, L.~D. (1986).
\newblock \emph{Fundamentals of Statistical Exponential Families: with
  Applications in Statistical Decision Theory}.
\newblock Institute of Mathematical Statistics, Hayward, CA.

\bibitem[Casella and Berger(2001)]{casella}
Casella, G. and Berger, R.~L. (2001).
\newblock \emph{Statistical Inference}, 2nd ed.
\newblock Pacific Grove, CA: Duxbury.

\bibitem[Cram\'{e}r(1946)]{cramer}
Cram\'{e}r, H. (1946).
\newblock \emph{Mathematical Methods of Statistics}.
\newblock Princeton University Press.

\bibitem[Fisher(1922)]{fisher}
Fisher, R.~A. (1922).
\newblock On the mathematical foundations of theoretical statistics.
\newblock \emph{Philosophical Transactions of the Royal Society of London.
    Series A}, \textbf{222}, 309--368.

\bibitem[Fristedt and Gray(1996)]{fristedt}
Fristedt, B.~E. and Gray, L.~F. (1996).
\newblock \emph{A Modern Approach to Probability Theory}.
\newblock Boston: Birkh\"{a}user.

\bibitem[Geyer(1990)]{thesis}
Geyer, C.~J. (1990).
\newblock Likelihood and exponential families.
\newblock Ph.D. thesis, University of Washington.
\newblock \url{http://purl.umn.edu/56330}.

\bibitem[Geyer(2009a)]{gdor}
Geyer, C.~J. (2009a).
\newblock Likelihood inference in exponential families and directions of
    recession.
\newblock \emph{Electronic Journal of Statistics}, \textbf{3}, 259--289
    (electronic).

\bibitem[Geyer(2009b)]{package}
Geyer, C.~J. (2009b).
\newblock R package aster: Aster models, version 0.7-7.
\newblock If the R statistical computing environment \citep{rcore}
has already been installed, then this package is installed
by the R command \verb@install.packages("aster")@ on any computer.

\bibitem[Geyer et al.(2007)Geyer, Wagenius and Shaw]{gws}
Geyer, C.~J., Wagenius, S. and Shaw, R.~G. (2007).
\newblock Aster models for life history analysis.
\newblock \emph{Biometrika}, \textbf{94}, 415--426.

\bibitem[Grosholz(submitted)]{grosholz}
Grosholz, E. (submitted).
\newblock Studying populations without molecular biology: Aster models and
    a new argument against reductionism.

\bibitem[H\'{a}jek(2007)]{hajek}
H\'{a}jek, A. (2007).
\newblock Interpretations of probability.
\newblock \emph{Stanford Encyclopedia of Philosophy}.
\newblock \url{http://plato.stanford.edu/}.

\bibitem[Jaynes(1978)]{jaynes}
Jaynes, E.~T. (1978).
\newblock Where do we stand on maximum entropy?
\newblock In \emph{The Maximum Entropy Formalism}, Ed. R.~D. Levine and
    M. Tribus, pp. 15-118.
\newblock MIT Press, Cambridge, MA.

\bibitem[Kolmogorov(1933)]{kolmogorov-foundations}
Kolmogorov, A.~N. (1933).
\newblock \emph{Grundbegriffe der Wahrscheinlichkeitsrechnung}.
\newblock Springer.
\newblock English translation (1950): \emph{Foundations of the
    theory of probability}. Chelsea.

\bibitem[Lande and Arnold(1983)]{la}
Lande, R., and Arnold, S.~J. 1983.
\newblock The measurement of selection on correlated characters.
\newblock \emph{Evolution}, \textbf{37}, 1210--1226.

\bibitem[Lauritzen(1996)]{lauritzen}
Lauritzen, S.~L. (1996).
\newblock \emph{Graphical Models}.
\newblock Oxford University Press, New York.

\bibitem[McCullagh and Nelder(1989)]{mccullagh}
McCullagh, P. and Nelder, J.~A. (1989).
\newblock \emph{Generalized Linear Models}, 2nd ed.
\newblock Chapman and Hall, London.

% \bibitem[Mitchell-Olds and Shaw(1987)]{ms}
% Mitchell-Olds, T., and Shaw, R.~G. 1987.
% \newblock Regression analysis of natural selection: Statistical inference
%     and biological interpretation.
% \newblock \emph{Evolution}, \textbf{41}, 1149--1161.

\bibitem[Oehlert(2000)]{oehlert}
Oehlert, G.~W. (2000).
\newblock \emph{A First Course in Design and Analysis of Experiments}.
\newblock New York: W.~H. Freeman.

\bibitem[R Development Core Team(2009)]{rcore}
R Development Core Team (2009).
\newblock R: A language and environment for statistical computing.
\newblock R Foundation for Statistical Computing, Vienna, Austria.
\newblock \url{http://www.R-project.org}.

\bibitem[Rockafellar and Wets(2004)]{raw}
Rockafellar, R.~T. and Wets, R. J.-B. (2004).
\newblock \emph{Variational Analysis}, corr.\ 2nd printing.
\newblock Berlin: Springer-Verlag.

\bibitem[Rudin(1986)]{rudin}
Rudin, W. (1986).
\newblock \emph{Real and Complex Analysis}, 3rd ed.
\newblock New York: McGraw-Hill.

\bibitem[Severini(2001)]{severini}
Severini, T.~A. (2001).
\newblock \emph{Likelihood Methods in Statistics}.
\newblock Oxford: Oxford University Press.

\bibitem[Shaw et al.(2008)Shaw, Geyer, Wagenius, Hangelbroek and
    Etterson]{aster2}
Shaw, R. G., Geyer, C. J., Wagenius, S., Hangelbroek, H. H., and
    Etterson, J. R. (2007)
\newblock Unifying life history analysis for inference of fitness
    and population growth.
\newblock \emph{American Naturalist}, \textbf{172}, E35--E47.

\bibitem[Shaw and Geyer(submitted)]{aster3}
Shaw, R. G. and Geyer, C. J. (submitted).
\newblock Inferring fitness landscapes.
\newblock Submitted to \emph{Evolution}.

\bibitem[Talbott(2008)]{talbott}
Talbott, W. (2008).
\newblock Bayesian Epistemology.
\newblock \emph{Stanford Encyclopedia of Philosophy}.
\newblock \url{http://plato.stanford.edu/}.

\bibitem[van der Vaart(2000)]{vdv}
van der Vaart, A.~W. (2000).
\newblock \emph{Asymptotic Statistics}
\newblock Cambridge University Press.

\bibitem[Weisberg(2005)]{weisberg}
Weisberg, S. (2005).
\newblock \emph{Applied Linear Regression}, 3rd ed.
\newblock Wiley, New York.

\end{thebibliography}

\end{document}

\begin{center} \LARGE REVISED DOWN TO HERE \end{center}

ISSUES

  DONE  1. "respecting constraints" needs a footnote about negative binomial
     and regular exponential families, but where?  Not appropriate in
     introduction before exponential families have been introduced.

  DONE  2. Need detailed work through of multivariate monotonicity, but where?
     Also not appropriate in introduction before exponential families have
     been introduced.

  DONE  3. Need more on submodels.  Curved manifolds for parameter spaces.

  SKIP IT  4. Rant about y = M beta + error ???

TO DO

  DONE  a. Explain Echinacea example in Shaw, et al. (2008).  Unconditional
     canonical parameterization loses track of effect of each variable,
     but that is the whole point!

PHILOSOPHY

  i. relationship between conditional and unconditional mean value parameter
     vectors peculiar to "predecessor is sample size" models.

  ii. multivariate monotone relationship between canonical and mean value
     parameters peculiar to exponential family models.

  iii. cumulant functions peculiar to exponential family models.  Hence
     mean and variance of canonical statistic from derivatives of cumulant
     functions.

  iv. relationship beween IID and exponential families
     peculiar to exponential family models.

  v. sufficient dimension reduction unique to exponential family models.

  vi. maximum entropy argument unique to exponential family models.

  vii. section on "philosophy of statistics" ???

  viii. "predecessor is sample size" interaction with IID and exponential
     families

  ix. no "missing" data, "structural zeros" handled automagically

\begin{center} \LARGE REVISED DOWN TO HERE \end{center}

The ``saying something'' about a parameter takes three forms: point estimates,
confidence intervals (and, more generally, confidence regions), and hypothesis
tests.  A point estimate of a parameter $\theta$ is a statistic (function of
the data that is not a function of the parameters) often denoted
$\hat{\theta}$.  Estimates are not parameters; $\hat{\theta}$ may be close
to the truth $\theta^*$ or far away.  Since $\hat{\theta}$ is a function
of random data, it is random and also described by a probability model
(the change-of-variable formula PDFs allows us to derive, in principle,
the PDF for $\hat{\theta}$ from the PDF $f_\theta$ of the data).
A confidence interval for a scalar parameter $\theta$ is an interval with
endpoints that are statistics.  Those endpoints being random, they are
described by a probability model derivable from $f_\theta$, so we can,
in principle, calculate the probability that the interval contains the truth
$\theta^*$, this probability, being derived from $f_\theta$ being a function
of $\theta$.

